---
title: "School Scores"
order: 60
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(GGally)
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

This dataset pertains to scores obtained by students in diverse subjects. Family Income is also part of this dataset. 

## Read the Data

```{r}
#| label: read-data
#| echo: false
#| message: false

scores <- read_csv(
  "../../../../../materials/Data/school_scores.csv")

```

{{< downloadthis ../../../../../materials/Data/school_scores.csv dname="school-scores-data" label="Download the School Scores Data" icon="database-fill-down" type="info" >}} 

## Inspect and Clean the Data

Hint: Use the `janitor` package here to clean up the variable names. Try to use the `big_camel` case name format for variables.

```{r}
#| echo: false
scores_clean <- scores %>% 
  janitor::clean_names(case = "big_camel")
glimpse(scores_clean)

```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Analyse the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```

## Plot the Data: All Subjects
```{r}
#| label: school-scores-plots-1
#| layout-ncol: 2
#| echo: false
## GGally Pairs Plot
scores_clean %>%
  select(contains("AverageGpa")) %>%
  rename_with(~ str_replace(., pattern = "AcademicSubjects", replacement = "")) %>%
  # remove the "AverageGpa" in the names
  rename_with(~ str_replace(., pattern = "AverageGpa", replacement = "")) -> scores_modified_1

scores_modified_1
#
scores_modified_1 %>%
  GGally::ggpairs(
    title = "Academic Scores in Diferent Subjects",
    subtitle = "Chart 1",
    progress = FALSE,
    switch = "both",
    lower = list(continuous = wrap(
      "smooth", alpha = 0.05, se = FALSE
    ))
  ) +
  theme_bw()

```

## Plot the Data: Maths vs Family Income
```{r}
#| label: school-scores-plots-2
#| layout-ncol: 2
#| echo: false
## Box Plots
scores_modified_2 <- scores_clean %>% 
  select(contains("KMath")) %>% 
  # remove the "FamilyIncome in the names
  rename_with(
  ~ str_replace(., pattern = "FamilyIncome",
              replacement = "")) %>% 
  # remove the "Math" in the names
   rename_with(
  ~ str_replace(., pattern = "Math",
              replacement = "")) %>%   
  pivot_longer(cols = everything(), 
               names_to = "names", 
               values_to = "values") %>% 
  mutate(names = base::factor(names, labels = unique(names),
                              levels = c("LessThan20K", 
                                         "Between20_40K",
                                         "Between40_60K",
                                         "Between60_80K",
                                         "Between80_100K",
                                         "MoreThan100K"),
                              ordered = TRUE)) 
scores_modified_2
# Plot this
scores_modified_2 %>% 
  gf_boxplot(values ~ reorder(names, values,FUN = median), 
             group = ~ names, fill = ~ names, 
             show.legend = FALSE,
            title = "Math Scores vs Family Income",
            subtitle = "Chart 2",
            xlab = "Income Class",
            ylab = "Scores in Math") %>% 
  gf_refine(coord_flip(), 
            scale_fill_grey(start = 0.8, end = 0.3)) %>% 
  gf_theme(theme_custom())

```

## Task and Discussion

Complete the Data Dictionary.
Select and Transform the variables as shown.
Create the graphs shown below and discuss the following questions:


-   Identify the type of charts
-   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
-   What activity might have been carried out to obtain the data graphed
    here? Provide some details.
-   What might have been the Hypothesis/Research Question to which the response was Chart #1?
- And Chart #2
- Write a 2-line story based on each of the graphs, describing your inference/surprise.

    

