---
title: "Correlation and Regression Explorations"
author: "Arvind V"
format: html
editor: visual
---

## Packages

```{r}
#| label: setup
library(tidyverse)
library(mosaic)
library(broom)
library(ggformula)
ggplot2::theme_set(new = theme_classic())

```

## Intro

I will work through and "unify" at least two things:

-   Hadley Wickham's chapter on modelling and his analysis of the linear
    model for the diamonds dataset
-   The diagnostic aspects of Linear Regression as detailed in Crawley's
    book

## Explorations into Diagnostic Plots

Let us create dependent y\* variables with different sorts of errors:

```{r}
x <- 0:300
en <- rnorm(301, mean = 0, sd = 5)
eu <- (runif(n = 301) -0.5) * 20
eb <- rnbinom(n = 301,prob = 0.3,size = 2)
eg <- rgamma(n = 301,shape = 1, rate = 1/x)
yn <- x + 10 + en
yu <- x + 10 + eu
yb <- x + 10 + eb
yg <- x + 10 + eg
data <- tibble(x, yn, yu, yb, yg)
data
```

### Normal Errors

```{r}
#| warning: false
lm_norm_aug <- lm(yn ~ x, data = data) %>% 
  augment()
lm_norm_aug %>% gf_point(.resid ~ .fitted)
lm_norm_aug %>% gf_qq(~ .resid) %>% gf_qqline()
```

### Uniform Errors

```{r}
#| warning: false
lm_unif_aug <- lm(yu ~ x, data = data) %>% 
  augment()
lm_unif_aug %>% gf_point(.resid ~ .fitted)
lm_unif_aug %>% gf_qq(~ .resid, distribution = stats::qnorm) %>% gf_qqline()
```


### Negative Binom Errors

```{r}
#| warning: false
lm_nbinom_aug <- lm(yb ~ x, data = data) %>% 
  augment()
lm_nbinom_aug %>% gf_point(.resid ~ .fitted)
lm_nbinom_aug %>% gf_qq(~ .resid, distribution = stats::qnorm) %>% gf_qqline()

```

### Gamma Errors

```{r}
#| warning: false
lm_gamm_aug <- lm(yg ~ x, data = data) %>% 
  augment()
lm_gamm_aug %>% gf_point(.resid ~ .fitted)
lm_gamm_aug %>% gf_qq(~ .resid, distribution = stats::qnorm) %>% gf_qqline()

```
