---
title: "\U0001F576 Science, Human Experience, Experiments, and Data"
subtitle: "Why do we visualize data"
author: "Arvind Venkatadri"
date: 2021-11-01
date-modified: "`r Sys.Date()`"
order: 10
format: html
#fig-format: retina
code-fold: true
code-summary: Show the Code
code-tools: true
code-line-numbers: true
code-overflow: wrap
code-copy: true
fig-align: center
reference-location: margin
cap-location: margin
highlight-style: dracula
df-print: paged
section-divs: true
number-sections: true
execute: 
  freeze: auto
tags: 
  - Scientific Inquiry
  - Experiments
  - Observations
  - Nature of Data
  - Experience
  - Measurement
---

## {{< fa folder-open >}} Slides and Tutorials

<a href="/content/slides/r-slides/nature-of-data/index.qmd">
<i class="fa-solid fa-file-powerpoint fa-2x"></i></a>     
<a href="/content/labs/r-labs/installation/index.qmd">
<i class="fa-solid fa-code fa-2xl"></i></a>

------------------------------------------------------------------------

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(blogdown)
library(tidyverse)
library(ggtext)

```

## {{< fa envelope title="An envelope" >}} Where does Data come from?

We will need to form a basic understanding of basic scientific
enterprise. Let us look at the slides.

## {{< fa chart-simple >}} Why Visualize?

-   We can digest information more easily when it is pictorial
-   Our [Working
    Memories](https://www.understood.org/en/articles/working-memory-what-it-is-and-how-it-works)
    are both *short-term* and *limited* in capacity. So a picture
    abstracts the details and presents us with an overall summary, an
    insight, or a story that is both easy to recall and easy on
    retention.\
-   Data Viz includes *shapes* that carry strong cultural memories and
    impressions for us. These cultural memories help us to use data viz
    in a *universal way* to appeal to a wide variety of audiences. (Do
    humans have a gene for geometry?)
-   It helps sift facts and mere statements: for example:

![Rape
Capital](/content/materials/images/rape-capital.png){fig-align="center"}

![What does Data
Reveal?](/content/materials/images/data-reveals-crime.png){fig-align="center"}

## {{< iconify mdi category-plus >}} What are Data Types?

{{< youtube dwFsRZv4oHA >}}

<br>

In more detail:

{{< youtube hZxnzfnt5v8 >}}

## {{< fa glasses >}} How do we Spot Data Variable Types? {#sec-data-types}

By asking questions!

::: column-page
+----------+----------+----------+----------+----------+
| Pronoun  | Answer   | Variable | Example  | What O   |
|          |          | / Scale  |          | p        |
|          |          |          |          | erations |
|          |          |          |          | ?        |
+==========+==========+==========+==========+==========+
| What,    | Name,    | Q u      | Name     | -        |
| Who,     | Place,   | a        |          |    Count |
| Where,   | Animal,  | litative |          |     of   |
| Whom,    | Thing    | or No    |          |          |
| Which    |          | m        |          |    Cases |
|          |          | inal\*\* |          | -   Mode |
+----------+----------+----------+----------+----------+
| How,     | A Manner | Q u      | So c i   | -        |
| What     | / Method | a        | o        |   Median |
| Kind,    | / Type   | litative | economic |          |
| What     | or       | or Or    | status,  | -   P e  |
| Sort     | A        | d        |          |     r    |
|          | ttribute | inal\*\* | E        | centiles |
|          | from a   |          | d        |          |
|          | List,    |          | ucation, |          |
|          | with     |          |          |          |
|          | list     |          | Income,  |          |
|          | items in |          |          |          |
|          | some     |          | S a t    |          |
|          | "order"  |          | i        |          |
|          | ( e.g    |          | sfaction |          |
|          | good,    |          |          |          |
|          | better,  |          |          |          |
|          | i        |          |          |          |
|          | mproved, |          |          |          |
|          | best...) |          |          |          |
+----------+----------+----------+----------+----------+
| How      | Q        | Q u a    | pH       | -   Mean |
| Many,    | u        | n        |          |          |
| How      | antities | titative | SAT      | -        |
| much,    | with     | or I n t | Score    | Standard |
| How      | Scale    | e        | (        |          |
| Heavy ,  |          | rval\*\* | 200-800) | D        |
| How few, | **Di f f |          |          | eviation |
| Seldom,  | e        |          | Credit   |          |
| Often,   | rences** |          | Score    |          |
| When     | are m e  |          | (        |          |
|          | a        |          | 300-850) |          |
|          | ningful, |          |          |          |
|          | but not  |          | Year of  |          |
|          | products |          | Starting |          |
|          | or       |          | College  |          |
|          | ratios   |          |          |          |
+----------+----------+----------+----------+----------+
| How      | Q u      | Q u a    | Weight,  | C o      |
| Many,    | a        | n        | Length,  | r        |
| How      | ntities, | titative | Height,  | relation |
| much,    | with     | or       |          |          |
| How      | Scale    | *        | T e m    | C o      |
| Heavy ,  | and a    | *Ratio** | p        | e        |
| How few, | Zero     |          | erature, | fficient |
| Seldom,  | Value.   |          | Enzyme   | of       |
| Often,   |          |          | A        | V        |
| When     | D i      |          | ctivity, | ariation |
|          | f        |          | Dosage   |          |
|          | ferences |          | Amount,  |          |
|          | and      |          | Reaction |          |
|          | Ratios   |          | Rate,    |          |
|          | /        |          |          |          |
|          | Products |          | C on c e |          |
|          | are m e  |          | n        |          |
|          | a        |          | tration, |          |
|          | ningful. |          | Pulse    |          |
|          | (e.g     |          | Rate,    |          |
|          | Weight ) |          | Survival |          |
|          |          |          | Time     |          |
+----------+----------+----------+----------+----------+

: Questions and Types of Variables
:::

As you go from `Qualitative` to `Quantitative` data types in the table,
I hope you can detect a movement from fuzzy groups/categories to more
and more crystallized numbers.

![Type of
Variables](/content/materials/images/Ratio-Interval-Ordinal-Nominal.PNG){fig-align="center"}

Each variable/scale can be subjected to the operations of the previous
group. In the words of [S.S.
Stevens](https://stats.idre.ucla.edu/other/mult-pkg/whatstat/what-is-the-difference-between-categorical-ordinal-and-interval-variables/)

> the basic operations needed to create each type of scale is
> cumulative: to an operation listed opposite a particular scale must be
> added all those operations preceding it.

## What Are the Parts of a Data Viz? {#sec-data-viz}

```{r echo=FALSE,message=FALSE}

g <- diamonds %>% slice_sample(n = 500) %>% ggplot(aes(x = carat, y = price, colour = cut)) + geom_point() +
  labs(title = "Plot Title", 
       subtitle = "Plot Subtitle",
       caption = "Plot Caption",
       x = "x-Axis Title",
       y = "y-Axis Title") + 
  guides(colour=guide_legend(title = "Plot Legend", 
                             colour = "orange")) + theme_bw() +
  theme(plot.title.position = "panel",
        plot.title = element_text(size = 20, 
                                  colour = "orange"),
        legend.title = element_text(colour = "orange", 
                                    size = 12),
        panel.background = element_rect(fill = "white"),
        plot.subtitle = element_text(size = 12, 
                                     colour = "orange"),
        plot.caption = element_text(size = 8, 
                                    colour = "orange"),
        axis.title.x = element_text(colour = "orange", 
                                    size = 12),
        axis.title.y = element_text(colour = "orange", 
                                    size = 12)
        )

annotation <- tibble(
   x = c(2,-0.01, 1.2),
   y = c(0, 10000, 17000),
   angle = c(0, 90, 0), 
   label = c("X-axis", "Y-axis", "Annotation")
)

# Add text
g + geom_text(data=annotation, aes( x=x, y=y, label=label, angle = angle),                 , 
           color="orange", 
           size=7 , fontface="bold" )
```

## How to pick a Data Viz?

Most Data Visualizations use one or more of the following geometric
attributes or aesthetics. These geometric aesthetics are used to
*represent* qualitative or quantitative variables from your data.

![Common Geometric Aesthetics in
Charts](/content/materials/images/common-aesthetics-1.png){fig-align="center"}

What does this mean? We can think of simple visualizations as
**combinations** of these aesthetics. Some examples:

::: column-body-outset
+-------------------+-------------------+-------------------+
| Aesthetic #1      | Aesthetic #2      | Shape             |
+===================+===================+===================+
| Position X =      | Position Y =      | Points/Circles    |
| Quant Variable    | Quant Variable    | with Fixed Size   |
+-------------------+-------------------+-------------------+
| Position X = Qual | Position Y =      | Columns           |
| Variable          | Count of Qual     |                   |
|                   | Variable          |                   |
+-------------------+-------------------+-------------------+
| Position X = Qual | Position Y = Qual | Rectangles, with  |
| Variable          | Variable          | *area*            |
|                   |                   | proportional to   |
|                   |                   | joint(X,Y)        |
|                   |                   | *count*           |
+-------------------+-------------------+-------------------+
| Position X = Qual | Position Y =      | Box + Whisker,    |
| Variable          | Rank-Ordered      | Box length        |
|                   | Quant Variable    | proportional to   |
|                   |                   | *Inter-Quartile   |
|                   |                   | Range*,           |
|                   |                   | whisker-length    |
|                   |                   | proportional to   |
|                   |                   | upper and lower   |
|                   |                   | quartile resp.    |
+-------------------+-------------------+-------------------+
| Position X =      | Position Y =      | Line and Area     |
| Quant Variable    | Quant Variable +  |                   |
|                   | Qual Variable     | Colour for Area   |
+-------------------+-------------------+-------------------+
|                   |                   |                   |
+-------------------+-------------------+-------------------+

: Geometries , Combinations, and Graphs
:::
