---
title: "Tutorial on Inference for Two Paired Means"
author: "Arvind Venkatadri"
date: 22/Nov/2022
date-modified: "`r Sys.Date()`"
---

## {{< iconify noto-v1 package >}} Setting up R Packages

```{r}
#| label: setup
#| include: true

library(tidyverse)
library(mosaic)

library(resampledata)

```


## {{< iconify pajamas issue-type-test-case >}} Case Study-1: IceCream!!

What is there to not like about icecreams!! Here is a dataset that has
data on Sugar and Calories between Vanilla and Chocolate icecreams,
across several brands of icecreams. Is this a sample of paired data? Let
us check:

### {{< iconify carbon chart-3d >}} Inspecting and Charting Data

```{r}
#| layout-nrow: 2
data("IceCream")
IceCream
inspect(IceCream)

```

Hmm...the data are about *calories*, *fat*, and *sugar* between two
flavours of icecream sold by **each** brand. There are 39 brands.

Let us plot the data first:

```{r}
#| layout-ncol: 3
IceCream %>% 
  gf_col(fct_reorder(Brand, VanillaCalories) ~ VanillaCalories, 
         fill = "red") %>% 
  gf_col(fct_reorder(Brand, VanillaCalories) ~ - ChocolateCalories, 
         fill = "green",
         xlab = "Calories", ylab = "Brand", 
         title = "Calories across Icecream Brands",
         subtitle = "Vanilla = Red, Green = Chocolate") %>% 
  gf_theme(theme_classic())

IceCream %>% 
  gf_col(fct_reorder(Brand, VanillaFat) ~ VanillaFat, 
         fill = "red") %>% 
  gf_col(fct_reorder(Brand, VanillaFat) ~ - ChocolateFat, 
         fill = "green",
         xlab = "Fat", ylab = "Brand", 
         title = "Calories across Icecream Brands",
         subtitle = "Vanilla = Red, Green = Chocolate") %>% 
  gf_theme(theme_classic())

IceCream %>% 
  gf_col(fct_reorder(Brand, VanillaSugar) ~ VanillaSugar, 
         fill = "red") %>% 
  gf_col(fct_reorder(Brand, VanillaSugar) ~ - ChocolateSugar, 
                  fill = "green",
         xlab = "Sugar", ylab = "Brand", 
         title = "Calories across Icecream Brands",
         subtitle = "Vanilla = Red, Green = Chocolate") %>% 
  gf_theme(theme_classic())

```

We may hypothesize that say, the fat content in the two flavours might
be similar on a **per brand** basis. That is, if say Baskin Robbins has
high `sugar` in the vanilla flavour, it is likely to have high sugar
also in its chocolate flavour.

Let us see what are the observed differences in the mean values of
*calories*, *sugar*, and *fat* across brands:

```{r observed-statistic-verizon}
IceCream %>% 
  mutate(diff_calories = VanillaCalories - ChocolateCalories,
         diff_fat = VanillaFat - ChocolateFat,
         diff_sugar = VanillaSugar - ChocolateSugar) %>% 
  summarise(mean_diff_calories = mean(diff_calories),
            mean_diff_fat = mean(diff_fat),
            mean_diff_sugar = mean(diff_sugar))

```

Hmm...while the numbers showing difference in means are quite different,
we need to perform tests to infer whether these difference are
statistically significant.

### {{< iconify academicons hypothesis >}} Hypothesis

How do we specify our Hypotheses? (Of course, there is more than one!)

Write the Null and Alternate hypotheses here.

### {{< iconify mdi null >}} {{< iconify tabler chart-histogram >}} Null Distribution Computations

How do we compute the NULL distributions, for each of the three
components of the ice creams, using pair-wise analysis?

```{r}
#| label:  null-distribution-by-permutation



```

## {{< iconify fluent-mdl2 decision-solid >}} Conclusions

So are there significant differences in sugar, fat, and calorie content
across the two flavours?

Is this conclusion different if you **don't** use paired-data, and just
treat the data as independent readings?
