---
title: <iconify-icon icon="icon-park-outline:sport" width="1.2em" height="1.2em"></iconify-icon> Heptathlon
order: 40
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(correlation)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

This is a dataset pertaining to scores of multiple athletes in the 7 events that make up the Heptathlon, modified for ease of analysis and plotting. 

## Data

```{r}
#| label: read-data
#| echo: true
#| message: false

library(HSAUR)
heptathlon

```

## Download the Modified data

Not Applicable!
```{r}
#| echo: false
#| eval: false

library(downloadthis)

job_gender_modified %>% 
   downloadthis::download_this(output_name = "job_gender_data", output_extension = ".csv", button_label = "Job and Gender Data", button_type = "default", icon = "fa fa-save")

# 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Analyse the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```

```{r}
#| label: corr-matrix
#| echo: false
hurdles_corr <- heptathlon %>%
  rownames_to_column(var = "name") %>% 
  dplyr::select(where(is.numeric)) %>% 
  # leave off hurdles to get all the remaining ones
  dplyr::select(-c(hurdles, score)) %>% 
  
  purrr::map(.x = .,
             .f = \(.x) cor.test(.x, heptathlon$hurdles)) %>% 
  
    # tidy up the cor.test outputs into a tidy data frame
  map_dfr(broom::tidy, .id = "predictor") 
hurdles_corr

```



## Plot the Data

```{r}
#| echo: false

hurdles_corr %>%
  
  gf_errorbar(conf.high + conf.low ~ reorder(predictor, estimate),
              #color = ~ estimate,
              width = 0.2,
              linewidth = ~ -log10(p.value)) %>% 
  
    # Reference line at zero correlation score
  gf_hline(yintercept = 0, color = "grey", linewidth = 2) %>% 
  
  # All correlation estimates as points
  gf_point(estimate ~ reorder(predictor, estimate), 
           shape = 21, size = 4,
           color = "white", fill = "black") %>% 
  

  
  # Themes,Titles, and Scales
  gf_labs(x = NULL, y = "Correlation with hurdles timing in Heptathlon",
          title = "Heptathlon Scores and Correlations",
          subtitle = "Which Events show Correlated Scores with Hurdles?",
          caption = "Significance = - log10(p.value)") %>% 
  
  gf_refine(
    scale_colour_gradient2("Correlation",low = "grey80", high = "grey0"),
    scale_linewidth_continuous("significance",
                                       range = c(0.2,4))) %>% 
  gf_refine(guides(linewidth = guide_legend(reverse = TRUE))) %>% 
  gf_theme(theme = theme_custom())

```

## Task and Discussion
Complete the Data Dictionary.
Create the graph shown and discuss the following questions:


-   Identify the type of charts
-   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
-   Which events in the 7-event heptathlon are most highly correlated
    with scores in hurdles?
-   If an athlete was a record holder in both high jump and hurdles,
    what would be your opinion about them? Justify based on the graph!
    

