---
title: <iconify-icon icon="fa:coffee"></iconify-icon> Coffee Flavours
subtitle: Coffee with Hansel and Gretel
order: 300
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggbump)
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

This dataset pertains to scores various types of coffees on parameters such as aroma, flavour, after-taste etc.  

::: callout-note
### Breadcrumbs
Since there are some interesting pre-processing actions required of data, and some choices to be made as well, I will leave some breadcrumbs, and some intermediate results, for you to look at and figure out the analysis/EDA path that you might take! You can then vary these at will after getting a measure of confidence!
:::

## Read the Data

```{r}
#| label: read-data
#| echo: false
#| message: false

coffee_ratings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-07/coffee_ratings.csv')
glimpse(coffee_ratings)
coffee_inspect <- inspect(coffee_ratings)
#coffee_inspect$quantitative

coffee_ratings %>%
  downloadthis::download_this(
    output_name = "coffee-ratings",
    output_extension = ".csv",
    button_label = "Download Coffee data",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```

## Inspect, Clean the Data

What are the non-numeric, or Qualitative variables here?

```{r}
#| label: inspect-categorical
#| echo: false
coffee_inspect$categorical
```

Look at the number of `levels` in those Qual variables!!Some are too many and some are so few... Suppose we count the data on the basis of a few?

```{r}
#| label: coffee-counts
#| echo: false
coffee_ratings %>% count(processing_method)
coffee_ratings %>% count(country_of_origin)
```

::: callout-note
### Breadcrumb 1
Why did I choose these Qual factors to count with?
:::


## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Research Question

::: callout-note
### Breadcrumb 1
Among the `country_of_origin` with the 5 highest *average* `total_cup_points`, how do the *average* ratings vary in ranks on the other coffee parameters?

Why this somewhat long-winded question? Why all this `average`stuff??

Why did I choose `country_of_origin`?Are there any other options? 
:::

## Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```

```{r}
#| label: coffee-long
#| echo: false
coffee_select <- coffee_ratings %>%
  select(
    country_of_origin,
    total_cup_points,
    aroma,
    flavor,
    aftertaste,
    acidity,
    body
  )

coffee_select

```
::: callout-note
### Breadcrumb 3
We have too much coffee here! We need to compress this data!

What??? Why? How? Where???
:::

```{r}
#| echo: false
# Take Averages of all ratings first
coffee_select %>%
  group_by(country_of_origin) %>%

# Reference: Ronak Shah
# https://stackoverflow.com/a/65046789
# 
  dplyr::summarise(across(
    .cols = where(is.numeric),
    .fns = list(mean = ~ mean(.x, na.rm = TRUE)),
    .names = "{.col}_{.fn}")) %>% 
  slice_max(order_by = total_cup_points_mean, n = 5) -> top_5_by_points

top_5_by_points

```
::: callout-note
### Breadcrumb 4
Where did all that coffee go???
Why are there only 5 rows in the data?
Why the names of the columns take on a surname, '_mean`??
:::

```{r}
#| label: rank-them
#| echo: false

top_5_by_points %>% 
  mutate(
    total_rank = dense_rank(desc(total_cup_points_mean)),
    aroma_rank = dense_rank(desc(aroma_mean)),
    aftertaste_rank = dense_rank(desc(aftertaste_mean)),
    body_rank = dense_rank(desc(body_mean)),
    acidity_rank = dense_rank(desc(acidity_mean))) %>% 
  select(country_of_origin,contains("rank")) -> top_5_by_ranks

top_5_by_ranks
```
::: callout-note
### Breadcrumb 5
What just happened? How did we convert those `mean` numbers to ranks?
:::

```{r}
#| label: coffee-plot
#| layout-ncol: 2
#| echo: false

top_5_by_ranks %>% 
  pivot_longer(cols = contains("rank"), 
               names_to = "coffee_parameter", 
               values_to = "ranks") %>% arrange(coffee_parameter) -> coffee_bump
coffee_bump
```

## Plot the Data
```{r}
#| echo: false
coffee_bump %>% 
ggplot(aes(x = coffee_parameter, y = ranks, 
           colour = country_of_origin, fill = country_of_origin,
           group = country_of_origin, shape = country_of_origin)) + 
  geom_point(size = 3) + 
  ggbump::geom_bump(linewidth = 1) + 
  scale_x_discrete("Country of Origin",
                   #breaks = c(1:5), 
    labels = c("acidity", "after taste", "aroma", "body","total")) +
  scale_shape_manual("Country of Origin", 
                     values = c(21:25)) +
  scale_y_reverse() +
  scale_color_brewer("Country of Origin", aesthetics = c("colour", "fill"), palette = "Set2") + 
  #scale_linetype_manual("Country of Origin", values = c(1,3,4,5,6)) +
  labs(title = "Rating Coffee from Around the World",
       y = "Rank") +
  theme_custom()

```

## Discussion

Complete the Data Dictionary.
Select and Transform the variables as shown.
Create the graphs shown below and discuss the following questions:


-   Identify the type of charts
-   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
-   What research activity might have been carried out to obtain the data graphed here? Provide some details.
-   What might have been the Hypothesis/Research Question to which the response was Chart?
- Write a 2-line story based on the chart, describing your inference/surprise.

    

