---
title: "\U0001F578 Networks"
subtitle: How things are connected
author: "Arvind Venkatadri"
date: 21/Nov/2022
date-modified: "`r Sys.Date()`"
order: 100
abstract: "Networks and Connections and what happens over them"
abstract-title: "Abstract"
format: html
#fig-format: retina
code-fold: false
code-summary: Show the Code
code-block-border-left: true
code-overflow: scroll
code-block-bg: "beige"
code-tools: true
code-line-numbers: true
code-copy: true
fig-align: center
reference-location: margin
cap-location: margin
highlight-style: tango
df-print: paged
section-divs: true
number-sections: true
execute: 
  freeze: auto
  cache: true
tags:
- Nodes and Edges
- Network Diagrams
- Tree Diagrams
- Arc Diagrams
---

# {{< fa folder-open >}} Slides and Tutorials

|     | <a href="./files/networks.ows"> <iconify-icon icon="fluent-emoji:orange-circle"></iconify-icon> Orange Tutorial</a> | <a href="https://r-for-artists.netlify.app/labs/07-networks/07-Grammar-of-Networks.html"><i class="fa-solid                fa-circle-nodes"></i> Networks in R</a>  |     |
|------------------|-------------------|------------------|------------------|

# Setting up R Packages

```{r}
#| label: setup
#| include: true
#| message: false
#| warning: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidygraph)
library(ggraph)
library(graphlayouts)
library(ggraph)
library(graphlayouts)
library(igraphdata)
library(sand)

```

# Introduction

Network graphs show relationships between entities: what sort they are,
how strong they are, and even of they change over time.

We will examine data structures pertaining both to the *entities* and
the *relationships* between them and look at the data *object* that can
combine these aspects together. Then we will see how these are plotted,
what the *structure* of the plot looks like. There are also *metrics*
that we can calculate for the network, based on its structure. We will
of course examine geometric metaphors that can represent various classes
of entities and their relationships.

Network graphs can be rendered both as static and interactive and we
will examine R packages that render both kinds of plots.

There is a another kind of structure: one that combines *spatial* and
*network* data in one. We will defer that for a future module !

## What kind Network graphs will we make?

Here is a network map of the characters in Victor Hugo's
`Les Miserables`:

<iframe width="100%" height="676" frameborder="0" src="https://observablehq.com/embed/@d3/force-directed-graph-canvas?cells=chart">

</iframe>

And this:

```{r}
#| echo: false
#| warning: false
data(karate)
set.seed(123)
l <- layout.graphopt(karate)
karate %>% 
  as_tbl_graph() %>% # to access nodes and edges in tidygraph
  
  # parameters to decorate the nodes
  activate(nodes) %>% 
  mutate(node_shape = if_else(V(karate)$label %in% c("H", "A"), 1, 2),
         node_size = 4*sqrt(igraph::degree(karate)),
         Faction = as_factor(Faction),
         node_shape = as_factor(node_shape)
         ) %>% 
  
  # parameters to decorate the edges
  activate(edges) %>%
  mutate(edge_color =
           case_when(
             edge_is_between(from = V(karate)$Faction == 1,
                             to = V(karate)$Faction == 2,
                             ignore_dir = TRUE) ~ 1L,
             edge_is_between(from = V(karate)$Faction == 1,
                             to = V(karate)$Faction == 1,
                             ignore_dir = TRUE) ~ 2L,
             edge_is_between(from = V(karate)$Faction == 2,
                             to = V(karate)$Faction == 2,
                             ignore_dir = TRUE) ~ 3L,
             TRUE ~ 4L),
         edge_color = as_factor(edge_color)) %>%
  
  # now plot it
  ggraph(., layout = l) +
  geom_edge_link0(aes(width = weight, 
                      color = edge_color)) + 
  geom_node_point(aes(size = node_size,
                      color = Faction,
                      shape = node_shape
                      )) +
  geom_node_text(aes(label = 
                       sub("Actor","",V(karate)$name)),repel = TRUE) + 
  scale_size(range = c(2, 10)) +
  scale_colour_manual(values = c("red", "dodgerblue")) +
  scale_shape_manual(values = c(15, 19)) +
  scale_edge_width(range = c(0.1, 4)) +
  scale_edge_colour_manual(values = c("yellow", "pink", "lightblue", "grey")) +
  theme_graph() + 
  theme(legend.position = "none")

```

## References

1.  [Hadley Wickham](http://hadley.nz/), [Danielle
    Navarro](https://djnavarro.net/), and [Thomas Lin
    Pedersen](https://www.data-imaginist.com/), *ggplot2: Elegant
    Graphics for Data Analysis*.
    <https://ggplot2-book.org/networks.html>

2.  Omar Lizardo and Isaac Jilbert, *Social Networks: An Introduction*.
    <https://bookdown.org/omarlizardo/_main/>

3.  Mark Hoffman, *Methods for Network Analysis.*
    <https://bookdown.org/markhoff/social_network_analysis/>

4.  *Statistical Analysis of Network Data with R, 2nd
    Edition*.<https://github.com/kolaczyk/sand>
