---
title: <iconify-icon icon="mingcute:hand-heart-line"></iconify-icon> Valentine's Day Spending
order: 90
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggprism)
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 16,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

This dataset pertains to spending on gifts by various people for Valentine's Day. This was part of the TidyTuesday Project for February 2024!

## Read the Data

```{r}
#| label: read-data
#| echo: true
#| message: false
gifts_age <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-02-13/gifts_age.csv')
glimpse(gifts_age)
#
gifts_age %>%
  downloadthis::download_this(
    output_name = "Valentines_Day_Gifts_dataset",
    output_extension = ".csv",
    button_label = "Download data",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )

```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::



## Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows
# Change data to factors etc.
# Set up Counts, histograms etc

```


## Research Question

::: callout-note
Write in!! Look at the Chart!
:::



## Plot the Data

```{r}
#| label: plot-gifts-data
#| echo: false
gifts_age %>% 
  pivot_longer(cols = -c(Age), 
               names_to = "SpendingOn", 
               values_to = "Amounts") %>% 
  gf_line(Amounts ~ Age, group = ~ SpendingOn, 
          linewidth = 1, color = ~ SpendingOn) %>% 
  gf_point( #linetype = ~ SpendingOn, 
          shape = ~ SpendingOn, size = 3) %>% 

  gf_labs(title = "Valentine's Day Spending over Age",
          x = "Age Group in Years",
          y = "Amounts Spent per Gift", 
          caption = "Tidy Tuesday 13-Feb-2024") %>% 
  gf_refine(scale_linetype_discrete("Items"),
            scale_colour_discrete("Items"),
            scale_shape_manual("Items", 
                values = c(0,1,2,4,5,6,8,10, 11)),
            scale_x_discrete(guide = "prism_bracket")) %>% 
  gf_theme(theme_custom())
  
```

## Task and Discussion

- Complete the Data Dictionary.
- Select and Transform the variables as shown.
- Create the graphs shown and discuss the following questions:
  -   Identify the type of charts
  -   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
  -   What research activity might have been carried out to obtain the data graphed here? Provide some details.
  - What pre-processing of the data was required to create the chart?
  -   What might be the Hypothesis / Research Question, based on the Chart?
  - Write a 2-line story based on the chart, describing your inference/surprise.


    

