---
title: "Hearing Loss in Children"
order: 110
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 16,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

Children are monitored for OME (fluid in the middle ear) over time. It is believed that they later suffer from "binaural hearing loss" (detecting sound amplitude and direction) after past episodes of OME during their childhood. The hearing-test is conducted multiple times, with a Test Signal embedded in noise played over audio loudspeakers. One loudspeaker has only Noise, and the other loudspeaker has the Test Signal in Noise. There are also two types of Test Signals: one is like noise itself and the other is distinct. In any test round, children are expected to orient themselves towards the appropriate loudspeaker and detect the presence of the Test Signal at varying levels of volume, with a passing success rate of 75% over multiple tests. 

This dataset is available on Vincent Arel-Bundock's dataset repository.

## Read the Data

```{r}
#| label: read-data
#| echo: true
#| message: false
ome <- read_csv("https://vincentarelbundock.github.io/Rdatasets/csv/MASS/OME.csv")
glimpse(ome)
ome
```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Research Question

::: callout-note
In hearing tests on people with varying levels of OME infection in their childhood, what is the effect of using distinct types of Test Signal on successful (face) orientation ?
:::


## Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows
# Rename Variables if needed
# Change data to factors etc.
# Set up Counts, histograms etc

```

```{r}
#| label: data-munging
#| echo: false
ome %>% 
  mutate(success_rate = Correct/Trials) %>% 
  mutate(ID = as_factor(ID), 
         Age = as_factor(Age),
         OME = base::factor(OME, levels = c("low", "high", "NA"), ordered = TRUE), 
         Noise = base::factor(Noise, labels = c("Test Signal is Noise-like", "Test Signal is Distinct"), levels = c("coherent", "incoherent")),
         Loud = as_factor(Loud)) -> ome_modified
ome_modified
```

## Plot the Data

```{r}
#| label: plot-ome-data
#| echo: false
ome_modified %>% 
    gf_boxplot(success_rate ~ OME) %>% 
  gf_facet_grid(~ Noise) %>% 
  #gf_hline(yintercept = 0.75, linewidth = 1, linetype = "dotted",colour = ~ "grey") %>% 
  gf_labs(title = "Hearing Test for Children with Otitis Media with Effusion (OME) Disease",,
          subtitle = "Test = Detection of  Test Signal in Noise\nPassing Score = 0.75",
          x = "Incidence of OME Disease over Childhood",
          y = "Successful Detection Score\n(over multiple trials)",
          caption = "Dataset from R package MASS by Venables and Ripley") %>% 
  gf_refine(scale_fill_grey(na.value = "grey90")) %>% 
  gf_theme(theme_custom())
  
```

## Task and Discussion

- Complete the Data Dictionary.
- Select and Transform the variables as shown.
- Create the graphs shown and discuss the following questions:
  -   Identify the type of charts
  -   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
  -   What research activity might have been carried out to obtain the data graphed here? Provide some details.
  - What pre-processing of the data was required to create the chart?
  - Write a 2-line story based on the chart, describing your inference/surprise. Is there something counter-intuitive (to a lay person) in the chart?


    

