---
title: "Ikea Furniture"
order: 10
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggridges)
#devtools::install_github('cttobin/ggthemr')
library(ggthemr)
library(ggsci)
# Theming
# ggthemr_reset()
# ggthemr::ggthemr("fresh")

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 


```


```{r}
library(ggthemr)
library(ggsci)
library(paletteer)
# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
##
## https://stackoverflow.com/questions/36476751/associate-a-color-palette-with-ggplot2-theme
## 
my_colours <- c("#fd7f6f", "#7eb0d5", "#b2e061", "#bd7ebe", "#ffb55a", "#ffee65", "#beb9db", "#fdcce5", "#8bd3c7")
my_pastels <- c("#66C5CC","#F6CF71","#F89C74","#DCB0F2","#87C55F","#9EB9F3","#FE88B1","#C9DB74","#8BE0A4","#B497E7","#D3B484","#B3B3B3")
my_greys <- c("#000000", "#333333", "#666666","#999999", "#cccccc")
my_vivids <- c("#E58606","#5D69B1","#52BCA3","#99C945","#CC61B0","#24796C","#DAA51B","#2F8AC4","#764E9F","#ED645A","#CC3A8E","#A5AA99")

my_bolds <- c("#7F3C8D","#11A579","#3969AC","#F2B701","#E73F74","#80BA5A","#E68310","#008695","#CF1C90","#f97b72","#4b4b8f","#A5AA99")

font <- "Roboto Condensed"
mytheme <- theme_classic(base_size = 14) + ### %+replace%    #replace elements we want to change
  
  theme(
    text = element_text(family = font),
    panel.grid.minor = element_blank(),
    #text elements
    plot.title = element_text(family = font,
      face = 'bold',
      hjust = 0, #left align
      #vjust = 2 #raise slightly
      margin = margin(0, 0, 10, 0)
    ),
    
    plot.subtitle = element_text(
      family = font,
      hjust = 0,
      margin = margin(2, 0, 5, 0)
    ),
    
    plot.caption = element_text(
      family = font,
      size = 8,
      hjust = 1
    ),
    #right align
    
    axis.title = element_text(#axis titles
      family = font, #font family
      size = 10), #font size
    axis.text = element_text(#axis text
        family = font, #axis family
        size = 8) #font size
    )
theme_av <- list(
  mytheme,
  scale_colour_manual(values = my_bolds, aesthetics = c("colour", "fill"))
)
  
```


## Introduction

This is a dataset pertaining to furniture prices at IKEA, modified for ease of analysis and plotting. 

## Data

```{r}
#| label: read-data
#| echo: false
#| message: false

library(tidytuesdayR)
tt <- tt_load("2020-11-03")
ikea <- tt$ikea 
ikea %>%
  select(-...1) %>%
  mutate(price_usd = 0.27 * price,
         short_description = str_squish(short_description)) %>%
  add_count(category, name = "category_total") %>%
  mutate(
    category = glue("{ category } ({ category_total })"),
    category = fct_reorder(category, price_usd)
  ) %>%
  filter(category_total >= 150) -> ikea_modified
ikea_modified

```

## Download the Modified data

```{r}
#| echo: false

library(downloadthis)

ikea_modified %>% 
   downloadthis::download_this(output_name = "ikea-modified", output_extension = ".csv", button_label = "Ikea furniture Data", button_type = "default", icon = "fa fa-save")

# 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
```
## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::


## Plot the Data

```{r}
#| echo: false
#
##
ikea_modified %>%
  ggplot(aes(price_usd, category, 
             fill = other_colors)) +
  geom_density_ridges(alpha = 0.75, aes(linetype = other_colors)) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    x = "Price (USD)",
    y = "",
    title = "How much do items in each IKEA category cost?",
    subtitle = "Are they available in other colors?",
    caption = "TidyTuesday Dataset for 03/11/2020"
  ) + theme_av


```

## Task and Discussion
Complete the Data Dictionary.
Create the graph shown and discuss the following questions:

-   What is the kind of plot used in the chart?
-   What variables have been used in the chart?
-   What can you say about the scale on X-axis?
-   What can you say about prices of items that are available in *single colour* versus those that are available in more than one colour?
-   What is a good hypothesis to interpret the double-humped nature of some of the curves?

