---
title: "Ikea Furniture"
order: 10
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   #size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   #size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

This is a dataset pertaining to furniture prices at IKEA, modified for ease of analysis and plotting. 

## Data

```{r}
#| label: read-data
#| echo: false
#| message: false

library(tidytuesdayR)
tt <- tt_load("2020-11-03")
ikea <- tt$ikea 
ikea %>%
  select(-...1) %>%
  mutate(price_usd = 0.27 * price,
         short_description = str_squish(short_description)) %>%
  add_count(category, name = "category_total") %>%
  mutate(
    category = glue("{ category } ({ category_total })"),
    category = fct_reorder(category, price_usd)
  ) %>%
  filter(category_total >= 150) -> ikea_modified
ikea_modified

```

## Download the Modified data

```{r}
#| echo: false

library(downloadthis)

ikea_modified %>% 
   downloadthis::download_this(output_name = "ikea-modified", output_extension = ".csv", button_label = "Ikea furniture Data", button_type = "default", icon = "fa fa-save")

# 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
```
## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::


## Plot the Data

```{r}
#| echo: false
ikea_modified %>% 
  ggplot(aes(price_usd, category, fill = other_colors)) +
  geom_density_ridges(alpha = .5, aes(linetype = other_colors)) +
  scale_x_log10(labels = dollar) +
  scale_fill_manual(values = c("grey10", "grey")) + 
  labs(x = "Price (USD)",
       y = "",
       title = "How much do items in each IKEA category cost?",
       subtitle = "Are they available in other colors?",
       caption = "TidyTuesday Dataset for 03/11/2020") +
  theme_custom()

```

## Task and Discussion
Complete the Data Dictionary.
Create the graph shown and discuss the following questions:

-   What is the kind of plot used in the chart?
-   What variables have been used in the chart?
-   What can you say about the scale on X-axis?
-   What can you say about prices of items that are available in *single colour* versus those that are available in more than one colour?
-   What is a good hypothesis to interpret the double-humped nature of some of the curves?

