---
title: <iconify-icon icon="icon-park-outline:proportional-scaling"></iconify-icon> Proportions
author: "Arvind V."
date: 27/Dec/2022
date-modified: '`r Sys.Date()`'
abstract: Types, Categories, and Counts
order: 40
categories:
- Proportions
- Frequency Tables
- Contingency Tables
- Numerical Data in Groups
- Margins
- Likert Scale data
- Bar Plots (for Contingency Tables)
- Mosaic Plots
- Balloon Plots
- Pie Charts
- Correspondence Analysis
bibliography: 
  - grateful-refs.bib
citation: true
editor: 
  markdown: 
    wrap: 72
webr:
  packages: ['readr', 'dplyr','mosaic', 'skimr', 'ggformula','GGally','kableExtra', 'vcd', 'vcdExtra', 'ggmosaic', 'resampledata', 'ggpubr']
---
:::: {.pa4}
::: {.athelas .ml0 .mt0 .pl4 .black-90 .bl .bw2 .b--blue}
[…"Thinking is difficult, that’s why most people judge."]{.f5 .f4-m .f3-l .lh-copy .measure .mt0}

[ --- C.G. Jung]{.f6 .ttu .tracked .fs-normal}
:::
::::

## {{< iconify noto-v1 package >}} Setting up R Packages {#sec-setting-up-r-packages}

```{r}
#| label: setup
#| include: true
#| message: false

library(tidyverse)
library(mosaic) # Our trusted friend
library(skimr)
library(vcd) # Michael Friendly's package, Visualizing Categorical Data
library(vcdExtra) # Categorical Data Sets
library(ggmosaic) # Mosaic Plots
library(resampledata) # More datasets

library(GGally) # Correlation Plots

library(ggpubr) # Colours, Themes and new geometries in ggplot
library(ca) # Correspondence Analysis, for use some day

## Making Tables
library(kableExtra) # html styled tables
library(gt)
library(tinytable)

```

```{r}
#| label: Extra Pedagogical Packages
#| echo: false
#| message: false

knitr::opts_chunk$set(tidy = TRUE)
library(checkdown)
library(epoxy)
# library(TeachHist)
# library(TeachingDemos)
library(grateful)

```

#### Plot Theme
```{r}
#| label: Plot Sizing and theming
#| code-fold: true
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),
      
      panel.grid.minor = element_blank(),    #strip minor gridlines
      
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   #size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   #size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())

```

## {{< iconify fe beginner >}} What graphs will we see today?

| Variable #1 | Variable #2 |       Chart Names       |                                            Chart Shape                                            |
|:---------------:|:---------------:|:-----------------:|:------------------:|
|    Qual     |    Qual     | Pies, and Mosaic Charts | {{< iconify teenyicons pie-chart-solid size=4x >}} {{< iconify carbon chart-marimekko size=4x >}} |

## {{< iconify tabler variable >}} What kind of Data Variables will we choose?

::: column-page-inset-right
```{r}
#| message: false
#| echo: false
#| warning: false
read_csv("../../../../../materials/Data/pronouns.csv") %>% 
  filter(No == "3") %>% 
  kbl() %>%
  kable_paper("hover", full_width = T)
  
```
:::

## {{< iconify openmoji japanese-symbol-for-beginner >}} Introduction

To recall, a categorical **variable** is one for which the possible
measured or assigned values consist of a discrete set of categories,
which may be `ordered` or `unordered`. Some typical examples are:

-   Gender, with categories "Male," "Female."
-   Marital status, with categories "Never married," "Married,"
    "Separated," "Divorced," "Widowed."
-   Fielding position (in ~~baseball~~ cricket), with categories
    "Slips,"Cover ","Mid-off "Deep Fine Leg", "Close-in", "Deep"...
-   Side effects (in a pharmacological study), with categories "None,"
    "Skin rash," "Sleep disorder," "Anxiety," . . ..
-   Political attitude, with categories "Left," "Center," "Right."
-   Party preference (in India), with categories "BJP" "Congress,"
    "AAP," "TMC"...
-   Treatment outcome, with categories "no improvement," "some
    improvement," or "marked improvement."
-   Age, with categories "0--9," "10--19," "20--29," "30--39," . . . .
-   Number of children, with categories 0, 1, 2, . . . .

As these examples suggest, categorical variables differ in the number of
categories: we often distinguish binary variables (or dichotomous
variables) such as Gender from those with more than two categories
(called *polytomous* variables).

## {{< iconify icons8 idea >}} Inspiration

::: {#fig-proportion-inspiration layout-ncol="2"}
![Obesity across the
World](../../../../../materials/images/VORO_Obesity-Rates_Site.jpg){#fig-pie-inspiration}

![Covid Deaths
<https://datatopics.worldbank.org/sdgatlas/goal-3-good-health-and-well-being?lang=en>](../../../../../materials/images/covid-deaths.png){#fig-mosaic-inspiration}

Depicting Proportions
:::

From @fig-pie-inspiration, it is seen that Egypt, Qatar, and the United
States are the only countries with a population greater than 1 million
on this list. Poor food habits are once again a factor, with some
cultural differences. In Egypt, high food inflation has pushed residents
to low-cost high-calorie meals. To combat food insecurity, the
government subsidizes bread, wheat flour, sugar and cooking oil, many of
which are the ingredients linked to weight gain. In Qatar, a country
with one of the highest per capita GDPs in the world, a genetic
predisposition towards obesity and sedentary lifestyles worsen the
impact of rich diets. And in the U.S., bigger portions are one of the
many reasons cited for rampant adult and child obesity. For example,
Americans ate 20% more calories in the year 2000 than they did in 1983.
They consume 195 lbs of meat annually compared to 138 lbs in 1953. And
their grain intake has increased 45% since 1970.

It’s worth noting however that this dataset is based on BMI values,
which do not fully account for body types with larger bone and muscle
mass.

From @fig-mosaic-inspiration, according to [World
Bank](https://datatopics.worldbank.org/sdgatlas/goal-3-good-health-and-well-being?lang=enS),
six countries (India, Russia, Indonesia, United States, Brazil, and
Mexico) accounted for over 60 percent of the total additional deaths in
the first two years of the pandemic.

## {{< iconify mdi food-processor-outline >}} How do these Chart(s) Work?

We saw with [Bar Charts](../20-BarPlots/index.qmd) that when we deal
with single Qual variables, we perform *counts* for each *level* of the
variable. For a single Qual variable, even with multiple levels ( e.g.
`Education Status`: High school, College, Post-Graduate, PhD), we can
count the observations as with Bar Charts and plot Pies.

We can also plot Pie Charts when the number of levels in a *single* Qual
variable are not too many. Almost always, a Bar chart is preferred. The
problem is that humans are pretty bad at reading angles. This ubiquitous
chart is much vilified in the industry and `bar charts` that we have
seen earlier, are viewed as better options. On the other hand, pie
charts are ubiquitous in design and business circles, and are very much
accepted! Do also read this spirited defense of pie charts here.
<https://speakingppt.com/why-tufte-is-flat-out-wrong-about-pie-charts/>

What if there are two Quals? Or even more? The answer is to take them
pair-wise, make all combinations of *levels* for both and calculate
counts for these. This is called a **Contingency Table**. Then we plot
that table. We'll see.

## {{< iconify mdi category-plus-outline >}} Categorical Data

From the `{vcd package}` vignette:

> The first thing you need to know is that categorical data can be
> represented in **three different forms in R**, and it is sometimes
> necessary to convert from one form to another, for carrying out
> statistical tests, fitting models or visualizing the results.

-   Case Data
-   Frequency Data
-   Cross-Tabular Count Data

Let us first see examples of each.

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
### Case Form

From Michael Friendly [Discrete Data Analysis and
Visualization](http://ddar.datavis.ca/pages/home) :

> In many circumstances, data is recorded on each individual or
> experimental unit. Data in this form is called case data, or data in
> case form. Containing *individual observations* with one or more
> **categorical** factors, used as classifying variables. The total
> number of observations is `nrow(X)`, and the number of variables is
> `ncol(X)`.

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
### {{< iconify la r-project >}} R

```{r}
#| label: arthritis-table

class(Arthritis)
# Tibble as HTML for presentation
Arthritis %>%  
  head(10) %>% 
  kbl(caption = "Case Form: Arthritis Treatments 
      and Effects<br> First 10 Observations", 
      centering = TRUE) %>%
  kable_classic_2(html_font = "Cambria", 
                  full_width = F) %>% 
  kable_styling(
    bootstrap_options = 
        c("striped", "hover", "responsive"))
```

### {{< iconify noto-v1 spider-web >}} web-r

```{webr-r}
#| label: case-form-data-webr
#| layout-ncol: 2

class(Arthritis)

# Tibble as HTML for presentation
Arthritis %>%  
  head(10) %>% 
  kbl(caption = "Arthritis Treatments and Effects<br> First 10 Observations", 
centering = TRUE) %>%
  kable_classic_2(html_font = "Cambria", 
full_width = F) %>% 
  kable_styling(bootstrap_options = 
c("striped", "hover", "responsive"))

```
:::

The `Arthritis` data set has three factors and two *integer* variables.
One of the three factors `Improved` is an **ordered factor**.

1)  ID
2)  Treatment: a factor; Placebo or Treated
3)  Sex: a factor, M / F
4)  Age: integer
5)  Improved: **Ordinal factor**; None \< Some \< Marked

Each row in the `Arthritis` dataset is a separate case or observation.

### Frequency Data Form

Data in frequency form has already been tabulated and *aggregated* by
**counting** over the (combinations of) categories of the table
variables. When the data are in **case form**, we can always trace any
observation back to its individual identifier or data record, since each
row is a unique observation or *case*; the reverse, with the **Frequency
Form** is rarely possible.

Frequency Data is usually a data frame, with columns of categorical
variables and at least one column containing `frequency` or `count`
information.

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
### {{< iconify la r-project >}} R

::: columns
::: column
```{r}
#| label: GSS-table
#| eval: false
str(GSS)

# Tibble as HTML for presentation
GSS %>%
  kbl(caption = "General Social Survey", 
centering = TRUE) %>%
  kable_classic_2(html_font = "Cambria", 
full_width = F) %>% 
  kable_styling(bootstrap_options = 
c("striped", "hover", "responsive"))

```
:::

::: column
```{r}
#| ref.label: GSS-table
#| echo: false

```
:::
:::

### {{< iconify noto-v1 spider-web >}} web-r

```{webr-r}

str(GSS)

# Tibble as HTML for presentation
GSS %>%
  kbl(caption = "General Social Survey",centering = TRUE) %>%
  kable_classic_2(html_font = "Cambria", full_width = F) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))

```
:::

Respondents in the GSS survey were classified by `sex` and `party`
identification. As can be seen, there is a **count** for every
combination of the two *categorical* variables, `sex` and `party`.

### Table form

Table Form Data can be a `matrix`, `array` or `table object`, whose
elements are the frequencies in an **n-way table**. The variable names
(factors) and their levels are given by `dimnames(X)`.

::: columns
::: column
```{r}
#| label:  Table-Form
#| eval: false

HairEyeColor
class(HairEyeColor)

```
:::

::: column
```{r}
#| ref.label: Table-Form
#| echo: false
```
:::

`HairEyeColor` is a "two-way" table, consisting of **two tables**, one
for `Sex = Female` and the other for `Sex = Male`. The total number of
observations is `sum(X)`. The number of dimensions of the table is
`length(dimnames(X))`, and the table sizes are given by
`sapply(dimnames(X), length)`. The data looks like a n-dimensional cube
and needs n-way tables to represent.

```{r}
#| label: exploring-tables
sum(HairEyeColor)
dimnames(HairEyeColor)
sapply(dimnames(HairEyeColor), length)

```

A good way to think of **tabular data** is to think of a Rubik's Cube.

![Rubik's Cube model for Multi-Table
Data](cube.jpg){fig-alt="Rubik's Cube model for Multi-Table Data"
fig-align="center" width="352"}

::: callout-tip
#### Rubik's Cube and Categorical Data Tables

Each of the edges is an *Ordinal Variable*, each segment represents a
*level* in the variable. So each face of the Cube represents *two*
ordinal variables. Any segment is at the intersection of two
(independent) levels of two variables, and the colour may be visualized
as a *count*. This array of counts on a face is a 2D or **2-Way Table**.
( More on this later )
:::

Since we can only print 2D tables, we hold one face in front and the
image we see is a 2-Way Table. Turning the Cube by 90 degrees gives us
another face with 2 variables, with one variable in common with the
previous face. If we consider two faces together, we get two 2-way
tables, effectively allowing us to contemplate 3 categorical variables.

Multiple 2-Way tables can be **flattened** into a single long table that
contains all counts for all combinations of categorical variables. This
can be visualized as "opening up" and laying flat the Rubik's cube, as
with a cardboard model of it.

```{r}
#| label: flattened-table
ftable(HairEyeColor)

```

Finally, we may need to convert the (multiple) tables into a data frame
or tibble:

```{r}
#| label: tables-to-data-frame
#| layout-ncol: 2
## Convert the two tables into a data frame
HairEyeColor %>% 
  as_tibble() 

# Tibble as HTML for presentation
HairEyeColor %>% 
  as_tibble() %>%  # Convert
  kbl(caption = "Hair Eye and Color") %>% 
  kable_classic_2(html_font = "Cambria", full_width = F) %>% 
  kable_styling(bootstrap_options = c("hover", "striped", "responsive"))

```
:::
:::

## {{< iconify carbon chart-3d >}} Simple Plots for Categorical Data

{{< video https://www.youtube.com/embed/7NhNeADL8fA >}}

We have already examined [Bar Charts](../20-BarPlots/index.qmd).

Pie Charts are discussed in [here.](../60-PartWhole/index.qmd)

These are both good for *single Qual variables*. Bars are more suited
when there are many *levels* and/or when there is more than one Qual
variable, as discussed earlier.

## {{< iconify icon-park-outline nested-arrows >}} Plotting Nested Proportions

When we want to visualize proportions based on Multiple Qual variables,
we are looking at what [Claus
Wilke](https://clauswilke.com/dataviz/nested-proportions.html) calls
*nested proportions*: groups within groups. Making counts with
combinations of levels for two Qual variables gives us a data structure
called a
[*Contingency*](https://www.pagebypagebooks.com/P_G_Wodehouse/Right_Ho_Jeeves/Chapter_13_p2.html)
*Table*, which we will use to build our plot for nested proportions. The
Statistical tests for Proportions ( the $\chi^2$ test ) also needs
Contingency Tables. The *Frequency Table* we encountered earlier is very
close to being a full-fledged Contingency Table; one only needs to add
the margin counts! So what is a Contingency Table?

### {{< iconify mdi rain-chance >}} {{< iconify grommet-icons table-add >}} Creating Contingency Tables {#sec-creating-contingency-tables}

From [Wolfram
Alpha](https://www.wolframalpha.com/input/?i=contingency+table):

> A contingency table, sometimes called a **two-way frequency table**,
> is a tabular mechanism with at least two rows and two columns used in
> statistics to present categorical data in terms of frequency counts.
> More precisely, an $r \times c$ contingency table shows the observed
> frequency of two variables the observed frequencies of which are
> arranged into $r$ rows and $c$ columns. The intersection of a row and
> a column of a contingency table is called a cell.

In this section we understand how to make Contingency Tables from each
of the three forms. We will use `vcd`, `mosaic` and the `tidyverse`
packages for our purposes. Then we will see how they can be visualized.

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
### Using mosaic

I think this is the **simplest and most elegant way** of obtaining
Contingency Tables:

```{r}
#| label: fig-sample-contingency-table
#| fig-cap: "Contingency Table"
#| echo: true
data("GSS2002", package = "resampledata")
gss2002 <- GSS2002 %>% 
  dplyr::select(Education, DeathPenalty) %>% 
  tidyr::drop_na(., c(Education, DeathPenalty))
##
mosaic::tally(DeathPenalty ~ Education, data = gss2002) %>% 
  addmargins() 

```

Plotting this as an HTML table, we get:

```{r}
#| label: fig-sample-contingency-table-picture
#| fig-cap: "Contingency Table Picture"
#| echo: false

data("GSS2002", package = "resampledata")
gss2002 <- GSS2002 %>% 
  dplyr::select(Education, DeathPenalty) %>% 
  tidyr::drop_na(., c(Education, DeathPenalty))

mosaic::tally(DeathPenalty ~ Education, data = gss2002) %>% 
  addmargins() %>% 
  as_tibble() %>% 
  pivot_wider(id_cols = DeathPenalty, names_from = Education, values_from = n) %>% 
  tt() %>% 
  style_tt(color = "grey") %>% 
  style_tt(i = 1, j = 2, color = "black", bold = TRUE, background = "yellow") %>% 
  style_tt(i = 2, j = 5, color = "black", bold = TRUE, background = "yellow") %>% 
   style_tt(i = 3, color = "black", bold = TRUE, background = "palegreen") 

```

How was this computed?

So $117$ is the number of people who `Left HS` *and* `Favor` the death
penalty, and $71$ is the count for `Bachelors` who `Oppose` the death
penalty. And so on.

### Using base R

```{r}
#| label: base-R-one-way-two-way

# One Way Table ( one variable )
table(Arthritis$Treatment) # Contingency Table

# 1-way Contingency Table
table(Arthritis$Treatment) %>% addmargins() # Contingency Table with margins

# 2-Way Contingency Tables
# Choosing Treatment and Improved
table(Arthritis$Treatment, Arthritis$Improved) %>% addmargins() 

# Choosing Treatment and Sex
table(Arthritis$Sex, Arthritis$Improved) %>% addmargins()

```

We can use `table()` ( and also `xtabs()` ) to generate
multi-dimensional tables too (More than 2-way) These will be printed out
as **a series of 2D tables**, one for each value/level of the "third"
parameter. We can then *flatten* this set of tables using **ftable()**
and add margins to convert into a Contingency Table:

```{r}
#| label: r-base-three-way-table
my_arth_table <- table(Arthritis$Treatment, Arthritis$Sex, Arthritis$Improved)
my_arth_table


# Now flatten 
ftable(my_arth_table) 
ftable(my_arth_table) %>% addmargins()

```

A bit strange that the column labels disappear in the `ftable` when
margins are added...maybe need to investigate the `FUN` argument to
`add_margins()`.

### Using vcd

The `vcd` ( Visualize Categorical Data ) package by Michael Friendly has
a convenient function to create Contingency Tables: `structable()`; this
function produces a 'flat' representation of a high-dimensional
contingency table constructed by recursive splits (similar to the
construction of mosaic charts/graphs). `structable` tends to render
**flat tables**, of the kind that can be thought of as a *"text
representation" of the `vcd::mosaic` plot*:

The arguments of `structable` are:

-   a formula $y + p \sim x + z$ which shows which variables are to be
    included as *columns* and *rows* respectively on a table;
-   a `data` argument, which can indicate a `data frame` from where the
    variables are drawn.

```{r}
#| label: vcd-data-frame-to-contingency-table
#| results: hold
#| layout-ncol: 2

arth_vcd <- vcd::structable(data = Arthritis, Treatment ~ Improved)
arth_vcd
class(arth_vcd)
arth_vcd %>% addmargins()

```

```{r}
#| layout-ncol: 2
#| results: hold
# With Margins
arth_vcd %>% as.matrix() %>% addmargins()

```

```{r}
#| label: vcd-multitable-to-contingency-table
#| layout-ncol: 2

# HairEyeColor is in multiple table form
HairEyeColor

# structable flattens these into one, as for a mosaic chart
vcd::structable(HairEyeColor)

# As tibble
vcd::structable(HairEyeColor) %>% as_tibble()

```

`UCBAdmissions` is already in Frequency Form i.e. a Contingency Table.
But it is a **set** of (two-way) Contingency Tables:

```{r}
#| layout-ncol: 2
UCBAdmissions
###
vcd::structable(UCBAdmissions)
###
structable(UCBAdmissions) %>% as.matrix() %>% addmargins()

```

::: callout-important
Note that `structable` does not permit the adding of margins directly;
it needs to be converted to a matrix for `addmargins()` to do its work.
:::


### Using tidyverse

So far these packages give Contingency Tables that are **easy to see for humans**; some of these structures are also capable being passed directly to commands such as `stats::chisq.test()` or `janitor::chisq.test()`.

Often we need Contingency Tables that are in `tibble` form, and we need to perform some data processing using `dplyr` to get there. Doing this with the `tidyverse` set of packages may seem [counter-intuitive and
long-winded](https://isha.sadhguru.org/en/wisdom/audio/the-story-of-ashtavakra-and-janaka), but the workflow is easily understandable.

First we develop the counts:

```{r}
#| label: tidyverse-for-Contingency-Tables
#| layout-ncol: 2
#| message: false
diamonds %>% count(cut)
diamonds %>% count(clarity)

diamonds %>% 
  group_by(cut, clarity) %>% 
  dplyr::summarise( count = n())

```

We need to have the individual levels of `cut` as rows and the
individual levels of `clarity` as columns. This means that we need to
pivot this from "long to wide"[^1] to obtain a Contingency Table:

```{r}
#| message: false
#| label: pivoting-long-to-wide
diamonds %>% 
  group_by(cut, clarity) %>% 
  dplyr::summarise( count = n()) %>% 
  
  pivot_wider(id_cols = cut, 
              names_from = clarity, 
              values_from = count) %>% 
  
  # Now add the row and column totals using the `janitor` package
  janitor::adorn_totals(where = c("row", "col")) %>%
  
  # Recover to tibble since janitor gives a "tabyl" format 
  # ( which can be useful too !)
  as_tibble() -> diamonds_ct
diamonds_ct

### Another Way 
diamonds %>% 
  group_by(cut, clarity) %>% 
  dplyr::summarise( count = n()) %>% 
  
  pivot_wider(id_cols = cut, 
              names_from = clarity, 
              values_from = count) %>% 
  
  # Now add the row and column totals using the `dplyr` package
  # From: https://stackoverflow.com/a/67885521
  mutate("row_totals"  = sum(across(where(is.integer)))) %>% 
  ungroup() %>% 
  add_row(cut = "col_total", summarize(., across(where(is.integer), sum)))

```

```{r}
#| include: false
#How about the data/tables we saw earlier?
# Convert to tibble
ftable(my_arth_table) %>% addmargins() %>% as_tibble()

# We can convert this to a tibble
arth_vcd %>% as_tibble()

# with margins? Ummm...
arth_vcd %>% as.matrix() %>% addmargins() %>% as_tibble()

tally(~ sex | substance, data = HELPrct, margins = TRUE) %>% as_tibble()

```
:::

[^1]: <https://tidyr.tidyverse.org/articles/pivot.html>

Now then, how does one plot a set of data that looks like this, a
matrix? No column is a single variable, nor is each row a single
observation, which is what we understand with the idea of [tidy
data](../05-NatureData/index.qmd#what-are-data-types).

### {{< iconify material-symbols auto-awesome-mosaic-outline-sharp >}} Mosaic Plots

The answer is provided in *the very shape of the data*: we plot this as
a set of *tiles*, where $$ \pmb{area~of~tile \sim count} $$ We
*recursively* partition off a (usually) square area into vertical and
horizontal pieces whose area is proportional to the count at a specific
combination of levels of the two Qual variables. So we might follow the
process as shown below:\

-   Take the [bottom row]{.bg-light-green .black} of per-column totals and create vertical
    rectangles with these widths\
-   Take the individual counts in the rows and partition each rectangle
    based in the counts in these rows.

::: {#fig-mosaic-plot-steps layout-ncol="2"}
![GSS Mosaic Chart Step
#1](../../../../../materials/images/Orange/gss-mosaic-chart-1.png){#fig-gss-mosaic-chart-step1}

![GSS Mosaic Chart Step
#2](../../../../../materials/images/Orange/gss-mosaic-chart-2.png){#fig-gss-mosaic-chart-step2}

Mosaic Chart for GSS Data
:::

The first split shows the various levels of `Education` and their counts
as widths. *Order is alphabetical!* This splitting corresponds to the
bottom ROW of the @fig-sample-contingency-table-picture. HS is clearly
the largest subgroup in `Education`.

In the second step, the columns from @fig-gss-mosaic-chart-step1 are
sliced horizontally into tiles, in proportion to the number of people in
each `Education` category/level who support/do not support
`DeathPenalty`. This is done in proportion to all the entries in each
COLUMN, giving us @fig-gss-mosaic-chart-step2.

Let us now make this plot with a variety of approaches.

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
#### Using vcd

The `vcd::mosaic()` function needs the data in *contingency table* form.
We already built one using `mosaic::tally()` and that is easily plotted:

```{r}

# Code used earlier
data("GSS2002", package = "resampledata")
gss2002 <- GSS2002 %>% 
  # select two categorical variables from the dataset
  dplyr::select(Education, DeathPenalty) %>% 
  drop_na(Education, DeathPenalty)
gss2002
# make a tally table
gss_table <- mosaic::tally(DeathPenalty ~ Education, data = gss2002)
gss_table %>% addmargins()

# gss_table is *not* a tibble, but a *table* object.
 
vcd::mosaic(gss_table, gp = shading_hsv,
main = "mosaic::tally() + vcd::mosaic()")

```

There is also a command within `vcd` itself to create a Contingency Table, `vcd::structable()`: 

```{r}
#| label:  mosaic-plot-using-vcd
#| layout-ncol: 2

arthritis_table <- vcd::structable(~ Treatment + Improved, 
                                   data = Arthritis)
arthritis_table

vcd::mosaic(arthritis_table, 
            gp = shading_max, 
            main = "Arthritis Treatment Dataset")


```


#### Using ggmosaic

`ggmosaic` takes a tibble with Qualitative variables, internally
computes the counts/table, and plots the mosaic plot:

::: columns
::: column
```{r}
#| label: gss2002
#| eval: false
gss2002
```
:::

::: column
```{r}
#| ref.label: gss2002
#| echo: false
```
:::
:::

::: columns
::: column
```{r}
#| label: ggmosaic-plot
#| eval: false
#library(ggmosaic)
#
# Set graph theme
theme_set(new = theme_custom())
##
ggplot(data = gss2002) +
  ggmosaic::geom_mosaic(
      aes(x = product(DeathPenalty, Education), 
          fill = DeathPenalty)) +
theme(legend.position = "top")

```
:::

::: column
```{r}
#| ref.label: ggmosaic-plot
#| echo: false
```
:::
:::

#### Using ggformula

This needs quite some work, to convert the Contingency Table into a
mosaic plot; perhaps not the most intuitive of methods either. This code
has been developed using this [Stackoverflow
post](https://stackoverflow.com/questions/19233365/how-to-create-a-marimekko-mosaic-plot-in-ggplot2).

```{r}
#| label: gss_summary_ggformula
#| fig-height: 6
#| fig-asp: 0.816
#| message: false
#| warning: false


# Reference
# https://stackoverflow.com/questions/19233365/how-to-create-a-marimekko-mosaic-plot-in-ggplot2


# Set graph theme
theme_set(new = theme_custom())

gss_summary <- gss2002 %>% 
  dplyr::group_by(Education, DeathPenalty) %>%
  dplyr::summarise(count = n()) %>% # This is good for a chisq test
  
  # Data is still grouped by `Education`
  # Add two more columns to facilitate mosaic Plot
  # These two columns are quite unusual...
  mutate(edu_count = sum(count), 
         edu_prop = count / sum(count)) %>%
  ungroup()
gss_summary

```

::: columns
::: column
```{r}
#| label: ggformula-marimekko
#| eval: false
# This works but is not very intuitive...
gf_col(edu_prop ~ Education, data = gss_summary,
       width = ~ edu_count, # Not typically used in a column chart
       fill = ~ DeathPenalty,
       stat = "identity", 
       position = "fill", 
       color = "black") %>% 
  
  gf_text(edu_prop ~ Education, 
          label = ~ scales::percent(edu_prop),
          position = position_stack(vjust = 0.5)) %>% 
  
  gf_facet_grid(~ Education, 
                scales = "free_x", 
                space = "free_x") %>% 
  
  gf_theme(scale_fill_manual(values = c("orangered", "palegreen3")))
  
```
:::

::: column
```{r}
#| ref.label: ggformula-marimekko
#| echo: false
```
:::
:::
:::

## Coloured Tiles: **Actual** and **Expected** Contingency Tables

We notice that the mosaic plots has coloured some tiles blue and some red. Why was this done? Consider the set of mosaic plots below:


::: column-body-outset-right
```{r}
#| label: fig-contingency-expected-actual-diff
#| fig-cap: 
#|   - "Actual Contingency Table"
#|   - "Expected Contingency Table"
#|   - "Tile-Wise Differences"
#| layout-ncol: 3

vcd::mosaic(arthritis_table, gp = shading_max, legend = FALSE)
vcd::mosaic(arthritis_table, type = "expected")
vcd::assoc(arthritis_table)

```
:::


From an inspection of these plots, we see the (**tile-wise**) difference between situations when Qualitative variables are *related* to that when
they *not related*. 

The graph on the left show the mosaic plot of the **actual** Contingency Table. 

The graph in the middle shows a similar but fictitious plot but with the cuts neatly horizontal or vertical. This mosaic would be what we would **expect**, if `Education` and the opinion on `Death Penalty` were independent!!

Clearly, there are differences in area of the corresponding tiles in the two mosaics, *actual* and *expected*, as shown in the graph on the right. Some differences are positive, and some negative. In the *actual mosaic*, @fig-contingency-expected-actual-diff-1, tiles with large positive differences are coloured blue, and those with large negative differences are coloured red.

The higher the absolute values of these differences, the greater the effect of one Qual on the other. More when we get into [Inference for Two Proportions.](../../../Inference/Modules/190-TwoProp/index.qmd)




::: {.content-hidden when-format="html"}
## Pearson Residuals in Mosaic Plots

The mosaic chart provides more additional information than does the
corresponding bar chart. The individual tiles are coloured based on the
value of the *Pearson Residual* (explained below). The Pearson residual
defines the (scaled) difference between the actual count in the cell,
with what might be expected if there was no association between the two
Qual variables. An uncoloured Mosaic Chart indicates that the residual differences are small and that there is no association.

So what are Pearson Residuals?

The mosaic chart is a visualization of the *obtained count* on which the
tile is constructed.

It is also possible to compute a per-cell *expected count*, if the
categorical variables are independent. The
*null hypothesis* is that the variables are independent. The test for
independence, as any inferential test, is based on comparing the
*observed counts* with the *expected counts*, under the null hypothesis.
So, what might the expected frequency of a cell be in cross-tabulation
table for cell $i,j$ given no relationship between the variables of
interest?

Represent the sum of row $i$ with $n_{+i}$, the sum of column $j$ with
$n_{j+}$, and the grand total of all the observations with $n$. And
[independence of variables means that their joint probability is the
product of their
probabilities](https://brilliant.org/wiki/probability-rule-of-product/).
Therefore, the Expected Cell Frequency/Count is given by:

$$
\begin{array}{lcl} ~Expected~Count~ e_{i,j} & = & P(Row~and~Column) * n\\
& = & P(row) \times P(column) * n\\
& = & \frac{rowSum}{n} *\frac{colSum}{n} \times n\\
& = &\frac{rowSum *colSum}{n}\\
\end{array}
$$


and therefore:

$$
e_{i,j} = \frac{(n_{+i})(n_{j+})}{n}
$$

Now, the **Pearson Residual** in each cell is equivalent to the
"z-score" of that cell.

Recall the [z-score](../22-Histograms/index.qmd#z-scores) idea: we
subtract the mean and divide by the std. deviation to get the z-score.
In the Contingency Table, we have **counts** which are usually modeled
as an (integer) *Poisson distribution*, for which **mean (i.e Expected
value) and variance are identical**. Thus we get the Pearson Residual
as:

$$
r_{i,j} = \frac{(Actual - Expected)}{\sqrt{\displaystyle Expected}}
$$ 

and therefore:


$$
r_{i,j} = \frac{(o_{i,j}- e_{i,j})}{\sqrt{\displaystyle e_{i,j}}}
$$

The comparison of what occurred to what is expected is based on their
difference, scaled by the square root of the expected, the **Pearson
Residual**.

The sum of all the squared Pearson residuals is the chi-square
statistic, χ2, upon which the inferential analysis follows. 

$$
χ2 = \sum_{i=1}^R\sum_{j=1}^C{r_{i,j}^2} 
$$ 


where R and C are number of rows and columns in the Contingency
Table, the levels in the two Qual variables.

We will treat the $X^2$ test in the Module on [Inference for Two
Proportions.](../../../Inference/Modules/190-TwoProp/index.qmd#observed-statistic-the-x2-metric).
See also the sub-section here below on *Actual and Expected Contingency
Tables*.

:::


## {{< iconify fluent people-list-28-filled >}} Dataset: Titanic

Banzai!!! That was quite some journey! Let us end it by quickly looking
at a sadly famous dataset:

```{r}
data("titanic", package = "ggmosaic")
titanic
```

There were 2201 passengers, as per this dataset.

### {{< iconify streamline dictionary-language-book-solid >}} Data Dictionary

::: callout-note
### Quantitative Data

None.
:::

::: callout-note
### Qualitative Data

-   `Survived`: (chr) yes or no
-   `Class`: (chr) Class of Travel, else "crew"
-   `Age`: (chr) Adult, Child
-   `Sex`: (chr) Male / Female.
:::

### {{< iconify material-symbols query-stats >}} Research Questions

::: callout-note
### Q.1. What is the dependence of `survived` upon `sex`?

::: columns
::: column
```{r}
#| label: vcd-titanic-1
#| eval: false
vcd::structable(Survived ~ Sex, data = titanic) %>% 
  vcd::mosaic(gp = shading_max)

```
:::

::: column
```{r}
#| ref.label: vcd-titanic-1
#| echo: false 
```
:::
:::

Note the huge imbalance in `survived` with `sex`: men have clearly
perished in larger numbers than women. Which is why the colouring by the
Pearson Residuals show large [*positive*
residuals]{style="color: blue;"} for men who died, and large [*negative*
residuals]{style="color: red;"} for women who died.

So sadly Jack is far more likely to have died than Rose.
:::

::: callout-note
### Q.2. How does `Survived` depend upon `Class`?

::: columns
::: column
```{r}
#| label: vcd-titanic-2
#| eval: false
vcd::structable(Survived ~ Class, data = titanic) %>% 
  vcd::mosaic(gp = shading_max)

```
:::

::: column
```{r}
#| ref.label: vcd-titanic-2
#| echo: false 
```
:::
:::

Crew has seen deaths in large numbers, as seen by the large [*negative*
residual]{style="color: red;"} for crew-survivals. [First Class
passengers have had speedy access to the
boats](https://youtu.be/kQ8WsD9AGX4) and have survived in larger
proportions than say second or third class. There is a large [*positive*
residual]{style="color: blue;"} for first-class survivals.

Rose travelled `first class` and Jack was `third class`. So again the
odds are stacked against him.
:::

## {{< iconify fluent-mdl2 balloons >}} Balloon Plots

There is another visualization of Categorical Data, called a Balloon
Plot. We will use the `housetasks` dataset from the package `ggpubr`.

### {{< iconify healthicons cleaning size = 2x >}} Dataset: Who Does the Housework?

```{r}
#| label: dataset-housetasks
#| 
housetasks <- read.delim(
  system.file("demo-data/housetasks.txt", 
              package = "ggpubr"), row.names = 1)
housetasks

inspect(housetasks)

```

We see that we have 13 observations.

::: callout-important
This data is already in Contingency Table form (without the margin
totals)!
:::

### {{< iconify streamline dictionary-language-book-solid >}} Data Dictionary

::: callout-note
### Quantitative Data

-   `Freq`: (int) No of times a task was carried out by specific people
:::

::: callout-note
### Qualitative Data

-   `Who`: (chr) Who carried out the task?
-   `Task`: (chr) Task? Which task? Can't you see I'm tired?
:::

::: columns
::: column
```{r}
#| label: balloon-plot-1
#| eval: false
# Set graph theme
theme_set(new = theme_custom())
##
ggpubr::ggballoonplot(housetasks, 
        fill = "value", 
        ggtheme = theme_pubr()) +
  scale_fill_viridis_c(option = "C") +
  labs(title = "A Balloon Plot for Categorical Data")

```
:::

::: column
```{r}
#| ref.label: balloon-plot-1
#| echo: false 
```
:::
:::

And repeat with the familiar `HairEyeColor` dataset:

::: columns
::: column
```{r}
#| label: HairEyeColor-tibble
#| eval: false
# Set graph theme
theme_set(new = theme_custom())

df <- as_tibble(HairEyeColor)
df
```
:::

::: column
```{r}
#| ref.label: HairEyeColor-tibble
#| echo: false 
```
:::
:::

::: columns
::: column
```{r}
#| label: balloon-plot-2
#| eval: false
ggballoonplot(df, x = "Hair", 
              y = "Eye", size = "n",
              fill = "n",
              ggtheme = theme_pubr()
              ) +
  scale_fill_viridis_c(option = "C") + 
  labs(title = "Balloon Plot")
```
:::

::: column
```{r}
#| ref.label: balloon-plot-2
#| echo: false 
```
:::
:::

::: columns
::: column
```{r}
#| label: balloon-plot-3
#| eval: false
# Balloon Plot with facetting
ggballoonplot(df, x = "Hair", 
              y = "Eye", size = "n",
              fill = "n", 
              facet.by = "Sex",
              ggtheme = theme_pubr()
              ) +
  scale_fill_viridis_c(option = "C") + 
  labs(title = "Balloon Plot with Facetting by Sex", 
       subtitle = "Hair and Eye Color")

```
:::

::: column
```{r}
#| ref.label: balloon-plot-3
#| echo: false 
```
:::
:::

Note the somewhat different syntax with `ggballoonplot`: the variable
names are enclosed in quotes.

Balloon Plots work because they use color and size aesthetics to
represent categories and counts respectively.

## {{< iconify mingcute thought-line >}} Wait, But Why?

-   We can detect *correlation* between Quant variables using the
    [scatter plots and regression lines](../30-Correlations/index.qmd)
-   And we can detect *association* between Qual variables using
    mosaics, sieves (which we did not see here, but is possible in R),
    and with balloon plots.
-   Your project primary research data may be pure Qualitative too, as
    with a Questionnaire / Survey instrument.
-   One such Qual variable therein will be your [target
    variable](https://orangedatamining.com/blog/machine-learning-jargon/)
-   You will need to justify whether the target variable is dependent
    upon the other Quals, and then to decide what to do about that.

## {{< iconify eos-icons ai >}} AI Generated Summary and Podcast

This excerpt from a course on Data Analysis using Metaphors focuses on the importance of understanding and visualizing *categorical data*. It discusses different ways to represent categorical data in R, including case data, frequency data, and cross-tabular count data. The text also explores various visualization techniques like bar plots, pie charts, mosaic plots, and balloon plots. It emphasizes the use of contingency tables for analyzing relationships between categorical variables, illustrating how to create them and visualize them using R packages. Additionally, the text delves into the concept of Pearson residuals, which help to identify associations between categorical variables and highlight deviations from independence.

<br><br>
<center>
```{r}
#| echo: false
embedr::embed_audio("../../../../../materials/audio/Categorical Data.wav")

``` 




## {{< iconify fluent-mdl2 decision-solid >}} Conclusion

How are the bar plots for categorical data different from histograms?
Why don't "regular" scatter plots simply work for Categorical data?
Discuss!

There are quite a few things we can do with Qualitative/Categorical
data:

1)  Make simple bar charts with colours and facetting
2)  Make Contingency Tables for a $X^2$-test
3)  Make Mosaic Plots to show how the categories stack up
4)  Make Balloon Charts as an alternative
5)  Then, draw your inferences and tell the story!

## {{< iconify openmoji person >}} Your Turn

1.  Take some of the categorical datasets from the `vcd` and `vcdExtra`
    packages and recreate the plots from this module. Go to
    <https://vincentarelbundock.github.io/Rdatasets/articles/data.html>
    and type "vcd" in the `search` box. You can directly load CSV files
    from there, using `read_csv("url-to-csv")`.

2.  Try the `housetasks` dataset that we used for Balloon Plots, to
    create a mosaic plot with Pearson Residuals.

::: callout-note
### [Clothing and Intelligence Rating of Children!!](https://vincentarelbundock.github.io/Rdatasets/doc/vcdExtra/Gilby.html)

Are well-dressed students actually smarter?

{{< downloadthis data/Gilby.csv dname="gilby" label="Download the Gilby Study dataset" icon="database-fill-down" type="info" >}}
:::

::: callout-note
### [Pre-marital Sex and Divorce](https://vincentarelbundock.github.io/Rdatasets/doc/vcd/PreSex.html).

{{< downloadthis data/Presex.csv dname="premarital" label="Download the pre- and extra-marital sex and divorce dataset" icon="database-fill-down" type="info" >}}
:::

::: callout-note
### [Are Emily and Greg More Employable Than Lakisha and Jamal?](https://vincentarelbundock.github.io/Rdatasets/doc/AER/ResumeNames.html)

Are first names a basis for racial discrimination, in the US?

This dataset was generated as part of a landmark research study done by
[Marianne Bertrand and Senthil
Mullainathan](https://sci-hub.se/10.1257/0002828042002561). Read the
description therein to really understand how you can ***prove
causality*** with a well-crafted research experiment.

{{< downloadthis data/ResumeNames.csv dname="ResumeNames" label="Download the Resume Name dataset" icon="database-fill-down" type="info" >}}
:::

## {{< iconify eos-icons ai >}} AI Generated Summary and Podcast

This module focuses on the importance of understanding and visualizing categorical data. It discusses different ways to represent categorical data in R, including *case data, frequency data, and cross-tabular count data*. The text also explores various visualization techniques like *bar plots, pie charts, mosaic plots, and balloon plots*. It emphasizes the use of *contingency tables* for analyzing relationships between categorical variables, illustrating how to create them and visualize them using R packages. Additionally, the text delves into the concept of *Pearson residuals*, which help to identify associations between categorical variables and highlight deviations from independence.

<br><br>
<center>
```{r}
#| echo: false
embedr::embed_audio("../../../../../materials/audio/change.wav")

``` 

## {{< iconify ooui references-ltr >}} References

1. Winston Chang (2024). *R Graphics Cookbook*. <https://r-graphics.org>

1.  Nice Chi-square interactive story at
    <https://statisticalstories.xyz/chi-square>

2.  Chittaranjan Andrade(July 22, 2015). *Understanding Relative Risk,
    Odds Ratio, and Related Terms: As Simple as It Can Get*.
    <https://www.psychiatrist.com/jcp/understanding-relative-risk-odds-ratio-related-terms/>

3.  Mine Cetinkaya-Rundel and Johanna Hardin. *An Introduction to Modern
    Statistics*, Chapter 4.
    <https://openintro-ims.netlify.app/explore-categorical.html>\

4.  Using the `strcplot` command from `vcd`,
    <https://cran.r-project.org/web/packages/vcd/vignettes/strucplot.pdf>\

5.  Creating Frequency Tables with `vcd`,
    <https://cran.r-project.org/web/packages/vcdExtra/vignettes/A_creating.html>\

6.  Creating mosaic plots with `vcd`,
    <https://cran.r-project.org/web/packages/vcdExtra/vignettes/D_mosaics.html>\

7.  Michael Friendly, *Corrgrams: Exploratory displays for correlation
    matrices*. The American Statistician August 19, 2002 (v1.5).
    <https://www.datavis.ca/papers/corrgram.pdf>\

8.  [Visualizing Categorical Data in
    R](https://homepage.divms.uiowa.edu/~luke/classes/STAT4580/catone.html)\

9.  H. Riedwyl & M. Schüpbach (1994), Parquet diagram to plot
    contingency tables. In F. Faulbaum (ed.), Softstat '93: Advances in
    Statistical Software, 293--299. Gustav Fischer, New York.

::: {#refs style="font-size: 60%;"}
###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("ggmosaic", "ggpubr", "janitor", "kableExtra", "resampledata", "sjPlot", "sjlabelled", "tinytable", "vcd",
           "vcdExtra")) %>%
  knitr::kable(format = "simple")

```
:::
