---
title: "Tutorial: Part of a Whole in R"
author: "Arvind V."
date: 12/Dec/2022
date-modified: "`r Sys.Date()`"
bibliography: 
  - grateful-refs.bib
citation: true
#suppress-bibliography: true
---

# Introduction

We will create Data Visualizations in R to show Parts of a Whole. As
always, we will consistently use the [**Project
Mosaic**](https://mosaic-web.org) ecosystem of packages in R (`mosaic`,
`mosaicData` and `ggformula`). Some specialized plots ( e.g. Fan Plots)
may require us to load other R Packages. These will be introduced
appropriately.

```{r, echo = FALSE, message = FALSE, include=TRUE}
knitr::opts_chunk$set(collapse = T, comment = "#>", echo = TRUE)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(tidyverse)
library(mosaic) # package for stats, simulations, and basic plots
library(mosaicData) # package containing datasets
library(ggformula) # package for professional looking plots, that use the formula interface from mosaic
library(NHANES) # survey data collected by the US National Center for Health Statistics (NCHS)
library(plotrix)

```


```{r}
#| label: Extra Pedagogical Packages
#| echo: false
#| message: false

library(checkdown)
library(epoxy)
library(grateful)

```

::: {#note .illustration style="background: beige"}
Recall the standard method for all commands from the `mosaic` package:

goal( y \~ x \| z, data = mydata, ...)
:::





## {{< iconify ooui references-rtl >}} References

::: {#refs style="font-size: 60%;"}

###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("ggridges", "NHANES", "TeachHist",
           "TeachingDemos")
) %>%
  knitr::kable(format = "simple")

```
:::
