---
title: "\U0001F409 Visualizing Survey Data"
author: "Arvind V."
date: 27/Dec/2022
date-modified: "`r Sys.Date()`"
abstract: Surveys, Questions, and Responses
order: 45
categories:
- Proportions
- Likert Scale data
bibliography: 
  - grateful-refs.bib
citation: true
editor: 
  markdown: 
    wrap: 72
webr:
  packages: ['readr', 'dplyr','mosaic', 'skimr', 'ggformula','GGally','kableExtra', 'vcd', 'vcdExtra', 'ggmosaic', 'resampledata', 'sjPlot', 'sjlabelled', 'ggpubr']
---

## {{< iconify noto-v1 package >}} Setting up R Packages {#sec-setting-up-r-packages}

```{r}
#| label: setup
#| include: true
#| message: false

library(tidyverse)
library(mosaic) # Our trusted friend
library(skimr)
library(vcd) # Michael Friendly's package, Visualizing Categorical Data
library(vcdExtra) # Categorical Data Sets
library(ggmosaic) # Mosaic Plots
library(resampledata) # More datasets

library(sjPlot) # Likert Scale Plots
library(sjlabelled) # Creating Labelled Data for Likert Plots

library(ggpubr) # Colours, Themes and new geometries in ggplot
library(ca) # Correspondence Analysis, for use some day

## Making Tables
library(kableExtra) # html styled tables

```

```{r}
#| label: Extra Pedagogical Packages
#| echo: false
#| message: false

library(checkdown)
library(epoxy)
# library(TeachHist)
# library(TeachingDemos)
library(grateful)

```

```{r}
#| label: Plot Sizing and theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   #size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   #size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())

```

## {{< iconify openmoji japanese-symbol-for-beginner >}} Introduction

In many business situations, we perform say *customer surveys* to get
[Likert Scale](https://www.simplypsychology.org/likert-scale.html) data,
where several respondents rate a product or a service on a scale of
`Very much like`, `somewhat like`, `neutral`, `Dislike` and
`Very much dislike,` for example.

## {{< iconify fluent likert-16-regular >}} Plots for Survey Data

How does this data look like, and how does one plot it? Let us consider a fictitious example, followed by a real world dataset.

## {{< iconify grommet-icons test >}} Case Study-1: A fictitious app Survey dataset

::: callout-note
### A fictitious QuickEZ app

We are a start-up that has an app called **QuickEZ** for delivery of
groceries. We conduct a survey of 200 people at a local store, with the
following questions,

1.  "Have your heard of the QuickEZ app?"
2.  "Do you use the QuickEZ app?"
3.  "Do you find it easy to use the QuickEZ app?"
4.  "Will you continue to use the QuickEZ app?"

where each questions is to be answered on a scale of : "always",
"often", "sometimes","never".
:::

Such data may look for example as follows:

```{r}
#| label: Demoing_labelled_data
#| layout-ncol: 2
#| column: body-outset-right
#| echo: false

#library(sjlabelled
set.seed(42)
score_labels = c("always", "often", "sometimes","never") # numerically 1:4
question_labels <- c("Have your heard of the QuickEZ app?",
                     "Do you use the QuickEZ app?",
                     "Do you find it easy to use the QuickEZ app?",
                     "Will you continue to use the QuickEZ app?")
quickez_data <- 
  # Create toy survey data
  # 200 responses to 5 questions
  # responses on Likert Scale
  # 1:4 = "never", "sometimes","often","always")
  # 
  tibble(q1 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.2, 0.2, 0.5, 0.1)),
         q2 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.3, 0.3, 0.3, 0.1)),
         q3 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.2, 0.1, 0.1, 0.6)),
         q4 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.4, 0.2, 0.1, 0.3))) %>%
  
  # Set VARIABLE labels
  sjlabelled::set_label(x = .,
                        label = question_labels) %>%
  
  # Now set VALUE labels
  sjlabelled::set_labels(x = ., labels = score_labels)

quickez_data %>%   head(10) %>% 
  kbl(caption = "First 10 Responses", centering = TRUE) %>%
  kable_classic_2(html_font = "Cambria", full_width = F) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))
str(quickez_data)

```

The columns here correspond to the 4 questions (q1-q4) and the rows
contain the 200 responses, which have been coded as (1:4). Such data is
also a form of *Categorical data* and we need to count and plot counts
for each of the survey questions. Such a plot is called a **Likert
plot** and it looks like this:

```{r}
#| label: fig-Likert-Plot-1
#| echo: false

# Set graph theme
theme_set(new = theme_custom())

plot_likert(quickez_data, title = "Summary of QuickEZ Questionnaire",
            #sort.frq = "" # Try "pos.desc" good news on top!
            reverse.colors = TRUE,
            show.prc.sign = TRUE
            ) 
```

Based on this chart, since it looks like about half the survey
respondents have not heard of our app, we need more publicity, and many
do not find it easy to use 😿, so we have serious re-design and user
testing to do !! But at least those who have managed to get past the
hurdles are stating they will continue to use the app, so it does the
job, but we can make it easier to use.

## {{< iconify grommet-icons test >}} Case Study-2: EUROFAM Survey dataset

Here is another example of Likert data from the *healthcare industry*.

`efc` is a German data set from a European study titled [**EUROFAM
study**](https://doi.org/10.1093/geront/48.6.752), on *family care of
older people*. Following a common protocol, data were collected from
national samples of approximately 1,000 family carers (i.e. caregivers)
per country and clustered into comparable subgroups to facilitate
cross-national analysis. The *research questions* in [**this EUROFAM
study**](https://doi.org/10.1093/geront/48.6.752) were:

> 1.  To what extent do family carers of older people use support
>     services or receive financial allowances across Europe? What kind
>     of supports and allowances do they mainly use?
>
> 2.  What are the main difficulties carers experience accessing the
>     services used? What prevents carers from accessing unused supports
>     that they need? What causes them to stop using still-needed
>     services?
>
> 3.  In order to improve support provision, what can be understood
>     about the service characteristics considered crucial by carers,
>     and how far are these needs met? and,
>
> 4.  Which channels or actors can provide the greatest help in
>     underpinning future policy efforts to improve access to
>     services/supports?

We will select the variables from the `efc` data set that related to
*coping* (on part of care-givers) and plot their responses after
inspecting them:

```{r}
#| label: efc_data
#| layout-nrow: 2
#| column: body-outset-right
#| echo: fenced
data(efc,package = "sjPlot")

efc %>% 
  select(dplyr::contains("cop")) %>% 
  head(20) 
##
efc %>% 
  select(dplyr::contains("cop")) %>% 
  str()

```

The `cop`ing related variables have responses on the Likert Scale
`(1,2,3,4)` which correspond to `(never, sometimes, often, always)`,
**and** each variable also has a `label` defining each variable. The
labels are actually ( and perhaps usually ) the *questions* in the
survey.

We can plot this data using the `plot_likert` function from package
`sjPlot`:

```{r}

# Set graph theme
theme_set(new = theme_custom())

efc %>% 
  select(dplyr::contains("cop")) %>% 
  sjPlot::plot_likert(title = "Caregiver Survey from EUROFAM") 

```

Many questions here have strong *negative* responses. This may indicate
that policy and publicity related efforts may be required.

::: callout-note
## Colours and Orientation in the Likert Plot

One could prefer (as I do) that "often" and "always" scores should be
toward the right and "sometimes" and "never" scores towards the left.
One can do this within the `plot_likert` command using:

`plot_likert(..., reverse.scale = TRUE)`

If you want the colours to be reversed, then...

`plot_likert(..., reverse.colors = TRUE)`

Try these options now in your Console! (Note the American spelling
`color`)
:::

## {{< iconify fluent-emoji label >}} Labelled Data

Note how the y-axis has been populated with **Survey Questions**: this
is an example of a *labelled dataset*, where not only do the variables
have *names* i.e. column names, but also have longish text **labels**
that add information to the data variables. The **data values** ( i.e
scores) in the columns is also **labelled** as per the the Likert scale
(`Like/Dislike/Strongly Dislike` OR `never/sometimes/often/always`) etc.
These **Likert scores** are usually a set of contiguous integers.

::: callout-note
## Variable Labels and Value Labels

**Variable label** is human readable description of the variable. R
supports rather long variable names and these names can contain even
spaces and punctuation but short variables names make coding easier.
Variable label can give a nice, long description of variable. With this
description it is easier to remember what those variable names refer
to.\

**Value labels** are similar to variable labels, but value labels are
descriptions of the values a variable can take. Labeling values means we
don't have to remember if `1=Extremely poor` and `7=Excellent` or
vice-versa. We can easily get dataset description and variables summary
with info function.
:::

Let us manually create one such dataset, since this is a common-enough
situation[^2] that we have survey data and then have to label the
variables and the values before plotting. We will use the R package
`sjlabelled` to label our data.[^3]. 

It is also possible to label the tibble, the columns, and the values in similar fashion using the `labelr` package.[^4]

[^2]: Piping Hot Data: Leveraging Labelled Data in R,
    https://www.pipinghotdata.com/posts/2020-12-23-leveraging-labelled-data-in-r/\>

[^3]: Label Support in
    R:<https://cran.r-project.org/web/packages/sjlabelled/index.html>
    
[^4]: Using the `labelr` package: <https://cran.r-project.org/web/packages/labelr/vignettes/labelr-introduction.html>

```{r}
#| label: Creating_labelled_data

#library(sjlabelled)

# Set graph theme
theme_set(new = theme_custom())

variable_labels <- c("Do you practice Analytics?",
                     "Do you code in R?",
                     "Have you published your R Code?",
                     "Do you use Quarto as your Workflow in R?",
                     "Will you use R at Work?")
value_labels = c("never", "sometimes","often","always") #numerically 1:4

my_survey_data <- 
  # Create toy survey data
  # 200 responses to 5 questions
  # responses on Likert Scale
  # 1:4 = "never", "sometimes","often","always")

  tibble(q1 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.2, 0.2, 0.5, 0.1)),
         q2 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.3, 0.3, 0.3, 0.1)),
         q3 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.2, 0.1, 0.1, 0.6)),
         q4 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.4, 0.2, 0.1, 0.3)),
         q5 = mosaic::sample(1:4, replace = TRUE, size = 200,
                             prob = c(0.1, 0.2, 0.5, 0.2))) %>%
  
  # Set VARIABLE labels
  sjlabelled::set_label(x = .,
                        label = variable_labels) %>%
  
  # Now set VALUE labels
  sjlabelled::set_labels(x = ., labels = value_labels)
###
head(my_survey_data, 6)
###
str(my_survey_data)

plot_likert(my_survey_data, 
            title = "Summary of Analytics Questionnaire",
            reverse.scale = TRUE,# Reverse score values on plot
            reverse.colors = FALSE, # let the colors be
            show.prc.sign = TRUE, # Show percentage sign
            legend.pos = "bottom")

```

It seems many people in the survey plan to use R at work!! And have
published R code as well. But Quarto seems to have mixed results! But of
course this is a toy dataset!!

So there we are with Survey data analysis and plots!

There are a few other plots with this type of data, which are useful in
very specialized circumstances. One example of this is the
`agreement plot` which captures the agreement between two (sets) of
evaluators, on ratings given on a shared ordinal scale to a set of
items. An example from the field of medical diagnosis is the opinions of
two specialists on a common set of patients. However, that is for a more
advanced course!

## {{< iconify fluent-mdl2 decision-solid >}} Conclusion

How are the Likert Plots for Survey data different from Bar Plots? Not very much inherently; we can view the Likert Charts as a set of stacked bar charts, based on Likert-scale response counts. At a pinch we can make a Likert Plot with vanilla bar graphs, but the elegance  and power of the packages `sjPlot` and `sjlabelled` is undeniable.

## {{< iconify openmoji person >}} Your Turn

1.  Take some of the categorical datasets from the `vcd` and `vcdExtra`
    packages and recreate the plots from this module. Go to <https://vincentarelbundock.github.io/Rdatasets/articles/data.html> and type "vcd" in the `search` box. You can directly load CSV files from there, using `read_csv("url-to-csv")`.

2. Including Edible Insects in our Diet!

{{< downloadthis data/Survey-on-Edible-Insects-likerted-copy.csv dname="edible-insects" label="Download the Edible Insects Dataset" icon="database-fill-down" type="info" >}}

There are several questions here for each "area" of preference for edible insects: experience, fear, concern for the environment, etc. Take all the columns marked as *average* as your data for your Likert Plot. 

## {{< iconify ooui references-ltr >}} References

1.  Mine Cetinkaya-Rundel and Johanna Hardin. *An Introduction to Modern
    Statistics*, Chapter 4.
    <https://openintro-ims.netlify.app/explore-categorical.html>\

2.  Using the `strcplot` command from `vcd`,
    <https://cran.r-project.org/web/packages/vcd/vignettes/strucplot.pdf>\

3.  Creating Frequency Tables with `vcd`,
    <https://cran.r-project.org/web/packages/vcdExtra/vignettes/A_creating.html>\

4.  Creating mosaic plots with `vcd`,
    <https://cran.r-project.org/web/packages/vcdExtra/vignettes/D_mosaics.html>\

5.  Michael Friendly, *Corrgrams: Exploratory displays for correlation
    matrices*. The American Statistician August 19, 2002 (v1.5).
    <https://www.datavis.ca/papers/corrgram.pdf>\

6.  [Visualizing Categorical Data in
    R](https://homepage.divms.uiowa.edu/~luke/classes/STAT4580/catone.html)\
    
7. H. Riedwyl & M. Schüpbach (1994), Parquet diagram to plot contingency tables. In F. Faulbaum (ed.), Softstat '93: Advances in Statistical Software, 293--299. Gustav Fischer, New York.


::: {#refs style="font-size: 60%;"}
###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("ggmosaic", "ggpubr", "janitor", "kableExtra", "resampledata", "sjPlot", "sjlabelled", "vcd",
           "vcdExtra")) %>%
  knitr::kable(format = "simple")

```
:::
