---
title: "A1"
author: "Arvind Venkatadri"
date: 03-March-2023
institute: "Dayanand Sagar University"
number-sections: false
slide-number: c/t
slide-tone: true
footer: "Assignment#1-IA-EMBA-T4-2023, Arvind Venkatadri"
scrollable: true
progress: true
touch: true
mouse-wheel: true
center: true
fig-height: 8
fig-width: 6
fig-align: center
fig-cap-location: bottom
section-divs: true
execute: 
  echo: false
  warning: false
  freeze: true
logo: DSU.png
format: revealjs
editor: visual
---

## Instructions

1.  Each Question in this Assignment is a chart.

2.  Each Chart is accompanied by a set of short questions.

3.  Your responses to these can be R-code, or text.

4.  Please number your answers as 1.a, 1.b, 1.c.....2.a, 2.b...on your
    Answer Sheet.

5.  All aboard? Let's go!

    ```{r}
    #| echo: false
    library(tidyverse)
    library(palmerpenguins)
    ```

## Question 1

::: columns
::: {#Q1 .column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() +  theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 1

::: columns
::: {.column width="40%"}
a.  What is the ggplot geometry used in this graph?
b.  What do the colours mean?
:::

::: {.column width="60%"}
```{r}
library(palmerpenguins)
ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g)) +  
  geom_density2d_filled() +  theme_minimal(base_size = 18) + theme(legend.position = "right") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 2

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 3

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 4

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 5

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 6

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 7

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 8

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 9

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 10

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 11

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 12

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 13

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 14

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::

## Question 15

::: columns
::: {.column width="50%"}
a.  List the variables are used in this graph?
b.  Identify their types. (Quantitative, Qualitative)
:::

::: {.column width="50%"}
```{r}
diamonds %>% ggplot(aes(y = price, x = carat, colour = cut)) + geom_point() + theme_minimal(base_size = 18) + theme(legend.position = "bottom") + guides(colour = guide_legend(override.aes = list(size = 4)))
```
:::
:::
