---
title: <iconify-icon icon="simple-icons:transportforlondon"></iconify-icon> <iconify-icon icon="hugeicons:water-pump"></iconify-icon>  William Farr's Observations on Cholera in London
order: 450
---


## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(GGally)
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

John Snow’s contention that cholera was principally spread by water was not accepted in the 1850s by the medical elite. The consequence of rejection was that hundreds in the UK continued to die.
William Farr, who founded the science of epidemiology, tried to examine if there were other causes that led to cholera. He had concluded that the available data not only supported miasma (spread via atmospheric vapours) but also indicated that there was an underlying ‘natural law’ linking infection with cholera inversely to elevation above high water.
The data is available on Vincent Arel-Bundock's website, and is part of the `HistData` package from Michael Friendly, UC Davis.

## Read the Data

```{r}
Cholera <- read_csv("https://vincentarelbundock.github.io/Rdatasets/csv/HistData/Cholera.csv")
Cholera
```


## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Research Question
::: callout-note
Write in! Look at the charts below!
:::

## Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```



## Plot the Data

```{r}
#| label: pairs-plot
#| echo: false
Cholera %>% 
  select(cholera_drate, poor_rate, elevation, pop_dens, persons_house, #water
         ) %>% 
  ggpairs(switch = "both",#aes(colour = water),
  # axis labels in more traditional locations(left and bottom)
  
  progress = FALSE,
  # no compute progress messages needed
  
  # Choose the diagonal graphs (always single variable! Think!)
  diag = list(continuous = "densityDiag"),
  # choosing histogram,not density
  
  # Choose lower triangle graphs, two-variable graphs
  lower = list(continuous = wrap("smooth", alpha = 0.3, se = FALSE)),
  
  title = "Plot 1: William Farr's Cholera Data",
  subtitle = "Correlation Plot"
) +
  
  theme_bw()
```

```{r}
#| echo: false
#| layout-ncol: 2
Cholera %>% 
  gf_point(cholera_drate ~ elevation, size = 3) %>% 
  gf_smooth(method = "lm",colour = "grey") %>% 
  gf_labs(x ="Elevation above high water mark (ft)",
              y ="Deaths from cholera in 1849 per 10,000",
          title = "Plot 2: Cholera Deaths vs Elevation above Thames High Water Mark") %>% 
  gf_refine(scale_y_continuous(limits = c(0, 250)),
            scale_colour_grey("Water Supply"),
            scale_shape("Water Supply")) %>% 
  gf_theme(theme = theme_custom())

Cholera %>% 
  gf_point(cholera_drate ~ elevation, shape = ~ water, color = ~ water, size = 3) %>% 
  gf_smooth(method = "lm") %>% 
  gf_labs(x ="Elevation above high water mark (ft)",
              y ="Deaths from cholera in 1849 per 10,000",
          title = "Plot 3: Cholera Deaths/n vs/nElevation above Thames High Water Mark") %>% 
  gf_refine(scale_y_continuous(limits = c(0, 250)),
            scale_colour_grey("Water Supply"),
            scale_shape("Water Supply")) %>% 
  gf_theme(theme = theme_custom())


```

## Tasks and Discussion

- Complete the Data Dictionary.
- Select and Transform the variables as needed.
- Look at Plot 1. Would you agree based on this chart that William Farr was right in believing that elevation was a good predictor for cholera deaths? Justify. 
- What is the nature of the relationship between Cholera Deaths and Elevation?
- Look at Plot 2. What kind of plot is it? What is the relationship here between Elevation and Cholera Death Rate? 
- Based on this graph, would you agree that Elevation is a predictor for Cholera Deaths? Justify. 
- Is the relationship you found between Cholera Deaths and Elevation also found in Plot 1? Justify.
- Look at Plot 3. Would you guess that there could be another predictor for Cholera Deaths? What could that Predictor be? Justify. 
    

