---
title: "California Transit Payments"
order: 200
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(correlation)
#
library(sjPlot)
library(sjlabelled)
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

This dataset is the result of a research study on payment options for people using public transit in California.  
The dataset is available on Dataset Dryad:

Pike, Susan (2022). *Transit payment preferences of unbanked passengers.* Dataset Dryad. <https://doi.org/10.25338/B8R04T>

and a brief 2-pager on the research methodology is [here](https://rosap.ntl.bts.gov/view/dot/61204/dot_61204_DS1.pdf).

## Read the Data

```{r}
#| label: read-data
#| echo: false
#| message: false

CalPayment <- read_csv("../../../../../materials/Data/doi_10_25338_B8R04T__v20220228/Transit_payment_preferences_of_unbanked_passengers.csv")

```

## Inspect and Clean the Data

```{r}
#| echo: false
#| eval: false
glimpse(CalPayment)
skimr::skim(CalPayment)
inspect(CalPayment)
names(CalPayment)
  
CalPayment2 <- CalPayment |> 
  dplyr::select(starts_with("phone")) %>% 
  mutate(across(everything(), ~ str_replace_all(., "1", "2"))) %>% 
  mutate(across(everything(), ~ str_replace_all(., "0", "1"))) %>% 
  mutate(across(everything(), as.numeric))

# ▢ Not having wifi access (1)
# ▢ Ways to add money (2)
# ▢ Identity concerns (3)
# ▢ Monthly fees (4)
# ▢ Knowing the balance (5)
# ▢ Other (please write in) (6)
 
# 
CalPayment2
CalPayment2 |> count(phone.balance)
CalPayment2 |> count(phone.wifi)
CalPayment2 |> count(phone.money)
CalPayment2 |> count(phone.identity)
CalPayment2 |> count(phone.fees)

variable_labels2 <- c("Wi_Fi access?",
                     "Ways to add money?",
                     "Identity Concerns?",
                     "Monthly Fees?",
                     "Knowing the balance?")


CalPayment2 <- CalPayment2 %>% 
  sjlabelled::set_label(x = ., label = variable_labels2) %>%
  sjlabelled::set_labels(x = ., labels = c("No", "Yes"))

str(CalPayment2)

```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Analyse the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```

```{r}
#| label: california-payment-data-1
#| echo: false
#| 
glimpse(CalPayment)
#skimr::skim(CalPayment)
# inspect(CalPayment)
# names(CalPayment)
```

Select and transform the data variables of interest:

```{r}
#| label: california-payment-data-2
#| echo: false
#| 
CalPayment2 <- CalPayment |> 
  dplyr::select(starts_with("phone")) %>% 
  mutate(across(everything(), ~ str_replace_all(., "1", "2"))) %>% 
  mutate(across(everything(), ~ str_replace_all(., "0", "1"))) %>% 
  mutate(across(everything(), as.numeric))

# ▢ Not having wifi access (1)
# ▢ Ways to add money (2)
# ▢ Identity concerns (3)
# ▢ Monthly fees (4)
# ▢ Knowing the balance (5)
# ▢ Other (please write in) (6)

glimpse(CalPayment2)

```


```{r}
#| echo: false
CalPayment2 |> count(phone.balance)
CalPayment2 |> count(phone.wifi)
CalPayment2 |> count(phone.money)
CalPayment2 |> count(phone.identity)
CalPayment2 |> count(phone.fees)

variable_labels2 <- c("Wi_Fi access?",
                     "Ways to add money?",
                     "Identity Concerns?",
                     "Monthly Fees?",
                     "Knowing the balance?")


CalPayment2 <- CalPayment2 %>% 
  sjlabelled::set_label(x = ., label = variable_labels2) %>%
  sjlabelled::set_labels(x = ., labels = c("No", "Yes"))

str(CalPayment2)
```


```{r}
#| label: california-payment-likert
#| echo: false

plot_likert(CalPayment2, reverse.colors = F, reverse.scale = F) +
  scale_fill_grey(start = 0.4, end = 0.8, labels = c("Yes", "No")) +
labs(title = "Challenges in Phone Payment\n in California Transit\n for Un-banked Passengers", caption = "Pike, Susan (2022). Transit payment preferences of unbanked passengers\n [Dataset]Dryad. https://doi.org/10.25338/B8R04T") +
  theme_custom()

```

## Task and Discussion
Complete the Data Dictionary.
Select and Transform the variables as shown.
Create the graph shown below and discuss the following questions:


-   Identify the type of charts
-   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
-   What activity might have been carried out to obtain the data graphed
    here? Provide some details.
-   What would be your recommendation to the Transport Company?
-   To the Phone Companies?
    

