---
title: "\U0001F4CE Correlations"
subtitle: "Correlations"
author: "Arvind Venkatadri"
date: 22/Nov/2022
date-modified: "`r Sys.Date()`"
order: 30
abstract: "How one variable changes with another"
format: html
#fig-format: retina
code-fold: false
code-summary: Show the Code
code-tools: true
code-line-numbers: true
code-overflow: scroll
code-copy: true
fig-align: center
reference-location: margin
cap-location: margin
highlight-style: tango
df-print: paged
section-divs: true
number-sections: true
execute: 
  freeze: auto
tags:
- Correlations  
- Scatter and Bubble Plots  
- 2D Density Plots  
- Heatmaps  
- Regression Lines  
---

# {{< fa folder-open >}} Slides and Tutorials

|                                                                    |                                                                                                                         |                                                                                                   |                                                                                        |
|---------------|--------------------|----------------|---------------|
| <a href="./files/correlations.qmd"><i class="fa-brands             
             fa-r-project"></i> R Tutorial</a>                       | <a href="./files/correlations.ows"> <iconify-icon icon="fluent-emoji:orange-circle"></iconify-icon> Orange Tutorial</a> | <a href="./files/correlations.rda"> <i class="fa-solid fa-person-rays"></i> Radiant Tutorial</a>  | <a href="./files/data/qdd-data.zip"> <i class="fa-solid fa-database"></i> Datasets</a> |


# Setup the Packages

```{r setup,message=FALSE}
library(tidyverse)
library(palmerpenguins)
library(patchwork)
library(ggformula)
library(mosaic)
library(GGally)
library(corrplot)

#library(grid)

```

# What graphs will we see today?

Some of the very basic and commonly used plots for data are:

-   Scatter Plot
-   Contour Plot
-   Correlogram
-   Heatmap

## What Correlation Charts can we plot with Numerical / Quantitative Data?

```{r charts, warning=FALSE,message=FALSE}
mpg <- mpg %>% mutate(drv= as_factor(drv))
p <- ggplot(mpg, alpha = 0.3) 
p1 <- gf_point(cty ~ hwy, fill = ~drv, colour = ~drv, data = mpg) %>% gf_smooth(method = "lm", title = "Scatter Plot with Correlation/Regression/Trend Line")
p2 <- gf_density_2d(cty~hwy, data = mpg, title = "2D Density Plot")
p1 + p2

```

```{r pairs-charts, warning=FALSE,message=FALSE}
p2 <- GGally::ggpairs(mpg, columns = c("hwy", "cty", "displ","cyl"), diag = list("densityDiag"), title = "Pairwise Correlations Plot")
p2

```

```{r cor-plot, warning=FALSE,message=FALSE}
library(corrplot)

mpg_num <- mpg %>% select(hwy, cty, displ, cyl)
mydata_cor <- cor(mpg_num)
corrplot(mydata_cor,method = "ellipse",type = "lower",main = "Correlogram")

# Heatmap with numbers
corrplot.mixed(mydata_cor, 
               lower = "color",
               upper = "number",
               bg = "wheat",
               tl.pos = "d",
               main = "Heatmap?")


```

## A Workflow in Orange

How does one calculate and visualize Correlations in Orange?

{{< video https://youtu.be/Bd7n9FdY3g0 >}}

Here is the help page on Correlations in Orange:

```{r,echo=FALSE}
knitr::include_url("https://orange3.readthedocs.io/projects/orange-visual-programming/en/latest/widgets/data/correlations.html")

```


## References

1.  Minimal R using `mosaic`.
    <https://cran.r-project.org/web/packages/mosaic/vignettes/MinimalRgg.pdf>
