---
title: <iconify-icon icon="fa6-solid:children"></iconify-icon> <iconify-icon icon="icon-park-solid:play-basketball"></iconify-icon> Children's Games
order: 80
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggbump)
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 16,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

Children in the ages of 6 to 7 years are asked if they want to play two games. This dataset pertains to their responses about the two games. The research is based on this paper:

Lin Bian et al. ,*Gender stereotypes about intellectual ability emerge early and influence children’s interests*. Science 355,389-391(2017).DOI:10.1126/science.aah6524. This very short and crisp paper is available [here.](https://www.science.org/doi/epdf/10.1126/science.aah6524)

## Read the Data
The data is part of the R package `openintro`. Yes, install it. 
From the help menu `?children_gender_stereo`:
 
>This data object is more unusual than most. It is a **list of 4 data frames**. The four data frames correspond to the data used in Studies 1-4 of the referenced paper, and these data frames each have variables (columns) that are explained below:

{{< pdf   files/OSF_Variable_summary.pdf >}}

```{r}
#| label: read-data
#| echo: true
#| message: false

library(openintro)
data("children_gender_stereo")
glimpse(children_gender_stereo)
```

```{r}
## Choosing, arbitrarily, the third game/third study
children_gender_stereo[[3]] ->games3
glimpse(games3)
```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::



## Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows
# Counts, histograms etc

```

```{r}
#| label: games3-munge
#| echo: false
#| message: false
#| warning: false
games3 %>%
  mutate(age = factor(age), 
         game = factor(game, 
                       labels = c("Smart Game", 
                                  "Try Hard Game"))) %>%
  group_by(age, game, gender) %>%
  dplyr::summarise(avg_interest = mean(interest)) -> games3_modified

games3_modified
```
## Research Question

::: callout-note
Is there a difference the average interest level between Boys and Girls for the two kinds of games, "Smart Game" and "Try Hard Game"? Does that lead to the inference of how children acquire gender stereotypes about play?
:::



## Plot the Data

```{r}
#| label: plot-games3
#| echo: false
games3_modified %>% 
  gf_col(avg_interest ~ age, group = ~ gender, 
                   fill = ~ gender, color = "black", 
                   position = "dodge") %>% 
gf_facet_wrap(vars(game)) %>% 
  gf_labs(title = "Interest in Play across Gender and Game Type",
          subtitle = "Difference from Baseline",
          x = "Age in Years", 
          y = "Average Percentage Interest in Playing Game",
          caption = "Bian et al., Science 355, 389–391 (2017) 27 January 2017") %>% 
  gf_theme(theme_custom())
  
```

## Task and Discussion

Complete the Data Dictionary.
Select and Transform the variables as shown.
Create the graphs shown below and discuss the following questions:


-   Identify the type of charts
-   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
-   What research activity might have been carried out to obtain the data graphed here? Provide some details.
-   Does the Chart answer the Hypothesis? Justify?
- Write a 2-line story based on the chart, describing your inference/surprise.


    

