---
title: <iconify-icon icon="noto:movie-camera" width="1.2em" height="1.2em"></iconify-icon> Movie Profits
order: 20
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   #size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   #size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

This is a dataset pertaining to movies and genres, modified for ease of analysis and plotting. 

## Data

```{r}
#| label: read-data
#| echo: false
#| message: false

library(tidytuesdayR)
movie_profit_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-10-23/movie_profit.csv")

# Data cleaning
movie_profit <- movie_profit_raw %>%
  select(-...1) %>%
  mutate(release_date = as.Date(parse_date_time(release_date, "%m!/%d/%Y"))) %>%
  filter(release_date < "2018-01-01") %>%
  arrange(desc(row_number())) %>%
  distinct(movie, release_date, .keep_all = TRUE) %>%
  mutate(distributor = fct_lump(distributor, 5)) %>%
  filter(worldwide_gross > 0) %>%
  mutate(profit_ratio = worldwide_gross / production_budget,
         decade = 10 * floor(year(release_date) / 10))
movie_profit

```

## Download the Modified data

```{r}
#| echo: false

library(downloadthis)

movie_profit %>% 
   downloadthis::download_this(output_name = "movie_profit", output_extension = ".csv", button_label = "Movie Profit Data", button_type = "default", icon = "fa fa-save")

# 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
```
## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::


## Plot the Data

```{r}
#| echo: false
movie_profit %>%
  group_by(genre, distributor) %>%
  summarize(median_profit_ratio = median(profit_ratio)) %>%
  arrange(desc(median_profit_ratio)) %>%
  mutate(genre = fct_reorder(genre, median_profit_ratio)) %>%
  ggplot(aes(genre, median_profit_ratio)) +
  geom_col() +
  labs(title = "Profits made by Film Distributors", subtitle = "Ratio of Profits to Budgets", caption = "Tidy Tuesday Oct 23, 2018") +
  scale_y_continuous(labels = function(x) paste0(x, "X")) +
  coord_flip() + 
  facet_wrap(~ distributor) +
  theme_custom()

```

## Task and Discussion
Complete the Data Dictionary.
Create the graph shown and discuss the following questions:

-   Identify the type of plot
-   What are the variables used to plot this graph?
-   If you were to invest in movie production ventures, which are the two best genres that you might decide to invest in?
-   Which R command might have been used to obtain the separate plots for each distributor?
-   If the original dataset had BUDGETS and PROFITS in separate columns, what preprocessing might have been done to achieve this plot?

