---
title: <iconify-icon icon="iconoir:sea-waves"></iconify-icon> <iconify-icon icon="mdi:weed"></iconify-icon>  Antarctic Sea ice
order: 400
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggbump)
library(ggprism)
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

```{r}
#| label: Plot-Sizing-and-theming
#| echo: false
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 20,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

Nine types of Seaweed were rated on different parameters and charted as shown below. 

::: callout-note
### Excel Data
The data is an excel sheet. Inspect it first in Excel and  decide which sheet you need, and which part of the data you need. There are multiple sheets!
Then use `readxl::read_xlsx(..)` to read it into R.
:::

## Read the Data

```{r}
#| label: read-data
#| echo: false
#| message: false

seaweed <- readxl::read_xlsx("../../../../../materials/Data/Seaweed_Farming_Marine_Permaculture.xls", sheet = "seaweed nutrition", range = "A3:R13")

seaweed %>%
  downloadthis::download_this(
    output_name = "seaweed-nutrition",
    output_extension = ".csv",
    button_label = "Download Seaweed Nutrition data",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```

## Inspect the Data

```{r}
#| label: inspect-categorical
#| echo: false
glimpse(seaweed)

```


## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Research Question
::: callout-note
Write in!
:::

## Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```

```{r}
#| label: coffee-long
#| echo: false
seaweed %>% 
  drop_na() %>% 
  dplyr::select("common_name" = `common name`, 
         "total_fibre" = `Total fiber`, Carbohydrates, 
         Calcium, Potassium, Zinc) %>% 
  mutate(fibre_rank = dense_rank(desc(total_fibre)),
         carbo_rank = dense_rank(desc(Carbohydrates)),
         calcium_rank = dense_rank(desc(Calcium)),
         potassium_rank = dense_rank(desc(Potassium)),
         zinc_rank = dense_rank(desc(Zinc))) %>% 
  select(common_name, contains("rank")) %>% 
  
  # Convert to Long Form
  pivot_longer(cols = contains("rank"), 
               names_to = "parameter", 
               values_to = "ranks") %>% 
  arrange(parameter) -> seaweed_bump

seaweed_bump

```


## Plot the Data
```{r}
#| echo: false
seaweed_bump %>%
  ggplot(
    aes(
      x = parameter,
      y = ranks,
      group = common_name,
      colour = common_name,
      shape = common_name
    )
  ) +
  ggbump::geom_bump(linewidth = 1) +
  
  scale_x_discrete(labels = c("Calcium", "Carbs", "Fibre", "Potassium", "Zinc"), guide = "prism_bracket") +
  geom_point(size = 3, fill = "black") +
  scale_shape_manual("Seaweed Type", values = c(0, 1, 2, 5, 7, 8, 15, 16, 18)) +
  scale_colour_discrete("Seaweed Type") +
  scale_y_reverse(breaks = c(1:9)) +
  scale_linetype_manual("Seaweed Type", values = c(1, 2, 3, 4, 5, 6, 1, 2, 3)) +
  
  labs(
    title = "Nutrients from Seaweed",
    caption = "Data from informationisbeautiful.net",
    x = "Nutrient",
    y = "Rank"
  ) +
  theme_custom()

```

## Tasks and Discussion

- Complete the Data Dictionary.
- Select and Transform the variables as shown.
- Create the graphs shown and discuss the following questions:
  -   Identify the type of charts
  -   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
  -   What research activity might have been carried out to obtain the data graphed here? Provide some details.
  -   What might have been the Hypothesis/Research Question to which the response was Chart?
  - Write a 2-line story based on the chart, describing your inference/surprise.
  - Based on the diagram, discuss which one an elderly person might try if they are deficient in calcium. If you were trying to avoid carbs, which seaweed sushi would you try?

    

