---
title: "Grain Transportation Cartels"
order: 120
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

#### Plot Theme
```{r}
#| label: Plot-Sizing-and-theming
#| code-fold: true
#| message: false
#| results: hide

# https://stackoverflow.com/questions/74491138/ggplot-custom-fonts-not-working-in-quarto

# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
### Ggplot Theme
### https://rpubs.com/mclaire19/ggplot2-custom-themes

theme_custom <- function(){ 
    font <- "Roboto Condensed"   #assign font family up front
    
    theme_classic(base_size = 14) %+replace%    #replace elements we want to change
    
    theme(
      panel.grid.minor = element_blank(),    #strip minor gridlines
      text = element_text(family = font),
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 16,               #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   #vjust = 2                #raise slightly
                   margin=margin(0,0,10,0)
),               
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,
                   margin=margin(2,0,5,0)
),               
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 8,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10                 #font size
),
      
      axis.text = element_text(              #axis text
                   family = font,            #axis family
                   size = 8)               #font size
    )
}

# Set graph theme
theme_set(new = theme_custom())
#
```

## Introduction

From: Robert H. Porter (1983). *A Study of Cartel Stability: The Joint Executive Committee, 1880-1886*. The Bell Journal of Economics, Vol. 14, No. 2 (Autumn, 1983), pp. 301-314:

The Joint Executive Committee (JEC) was a [cartel](https://en.wikipedia.org/wiki/Cartel) (of railroad firms) which controlled eastbound freight shipments from Chicago to the Atlantic seaboard in the 1880's. While different railroad firms in the JEC shipped grain to different port cities (for example, Baltimore and New York), most of the wheat handled by the cartel was subsequently exported overseas, and the rates charged by different firms (were) adjusted to compensate for differences in ocean shipping rates. 

Prices, rather than quantity, has typically been thought to be the strategic variable of firms in the rail-freight industry. Total
demand was quite variable, and so the actual market share of any particular railroad firm would depend on both the prices charged by all the firms as well as unpredictable (random) forces. Price wars were not random, but precipitated by periods of slackened demand, which were **presumably unpredictable**, at least to some extent. 

On the other hand, the **predictable** fluctuations in demand that resulted from the annual opening and closing of the Great Lakes (Superior / Michigan / Huron / Ontario / Erie ) to shipping (because they were frozen in winter), which determined the degree of outside competition, did not disrupt industry conduct. Rather, rates adjusted systematically with the lake navigation season. 

This dataset is available on Vincent Arel-Bundock's dataset repository, and is part of the R package `AER` (Applied Econometrics in R).

## Read the Data

```{r}
#| label: read-data
#| echo: true
#| message: false
cartelstability <- read_csv("https://vincentarelbundock.github.io/Rdatasets/csv/AER/CartelStability.csv")
cartelstability
glimpse(cartelstability)
```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Research Question

::: callout-note
How do prices for per-tonne grain transport vary based on whether the cartel is working or not? Does this depend upon whether it is summer time or winter time? Why?
:::


## Inspect/Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows
# Rename Variables if needed
# Change data to factors etc.
# Set up Counts, histograms etc

```

```{r}
#| label: data-munging
#| echo: false
cartel_clean <- cartelstability %>%
  mutate(season = stringr::str_squish(season)) %>%
  mutate(season = 
           factor(
    season,
     levels =
      c(
        "Jan 1 - Jan 28",
        "Jan 29 - Feb 25",
        "Feb 26 - Mar 26" ,
        "Mar 27 - Apr 23",
        "Apr 24 - May 21",
        "May 22 - Jun 18",
        "Jun 19 - Jul 16",
        "Jul 17 - Aug 13",
        "Aug 14 - Sep 10",
        "Sep 11 - Oct 8",
        "Oct 9 - Nov 5",
        "Nov 6 - Dec 3",
        "Dec 4 - Dec 31"),
    labels =
      c(
        "Jan 1 - Jan 28",
        "Jan 29 - Feb 25",
        "Feb 26 - Mar 26" ,
        "Mar 27 - Apr 23",
        "Apr 24 - May 21",
        "May 22 - Jun 18",
        "Jun 19 - Jul 16",
        "Jul 17 - Aug 13",
        "Aug 14 - Sep 10",
        "Sep 11 - Oct 8",
        "Oct 9 - Nov 5",
        "Nov 6 - Dec 3",
        "Dec 4 - Dec 31"),
    ordered = TRUE)) %>%
  mutate(ice = factor(ice, levels = c("yes", "no"),
                      labels = c("Lakes are Frozen", 
                                  "Lakes are not Frozen"))) %>% 
  mutate(cartel = factor(cartel, levels = c("yes", "no"),
                         labels = c("Cartel Operational", 
                                    "Cartel not Operational"))) 
cartel_clean
```
Some summarizing...

```{r}
#| echo: false
cartel_clean %>% 
  group_by(season, cartel, ice) %>% 
  summarise(avg_price = mean(price), 
            mean_tonnage = mean(quantity)) -> sum_cartel
sum_cartel
```

## Plot the Data

```{r}
#| label: plot-data
#| echo: false
sum_cartel %>% 
  ggplot(aes(season, avg_price,
             colour = cartel, 
             group = cartel)) +
  geom_point() +  geom_line() +
  facet_grid(cols = vars(ice), scales = "free_y") + 
  labs(y = "Avg. Price Index per Ton", 
       x = "Season (in 4 week chunks)",
       title = "Price Index per Ton of Grain Transported by Railway Firms",
       subtitle = "Railways Firms were part of JEC Cartel",
       caption = "Robert H. Porter.A Study of Cartel Stability:The Joint Executive Committee, 1880-1886.\n The Bell Journal of Economics. Vol. 14, No. 2, August 1983") + 
  scale_color_manual("", values = c("black", "grey70")) + 
  theme_custom() + 
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1),
        legend.position = "bottom") 
  
```

## Task and Discussion

- Complete the Data Dictionary.
- Select and Transform the variables as shown.
- Create the graphs shown and discuss the following questions:
  -   Identify the type of charts
  -   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
  -   What research activity might have been carried out to obtain the data graphed here? Provide some details.
  - What pre-processing of the data was required to create the chart?
  - Explain what happens when it is stated "cartel is working" and "cartel is not working". 
  - How do prices for per-tonne grain transport vary based on whether the cartel is working or not? Does this depend upon whether it is summer time or winter time? Why?
  - Is the cartel beneficial for customers of the JEC? What would be their behaviour based on whether the cartel was operational or not?
  


    

