---
title: "EDA: Exploring Static  Graphs for Distributions in R"
author: "Arvind V."
date: 19/Nov/2022
date-modified: "`r Sys.Date()`"
abstract: How to ask Questions of your Data and answer with Stats, and interactive Bar charts and Histograms with the `ggformula` package

---

## {{< iconify noto-v1 package >}} Setup the Packages

```{r}
#| label: setup
#| include: true
#| warning: false
#| message: false

options(digits = 3)
library(tidyverse)
library(mosaic) # package for stats, simulations, and basic plots
library(mosaicData) # package containing datasets
library(ggformula) # package for professional looking plots, that use the formula interface from mosaic
library(skimr) # Summary statistics about variables in data frames
library(NHANES) # survey data collected by the US National Center for Health Statistics (NCHS)

```

## {{< iconify fxemoji japanesesymbolforbeginner >}} Introduction

We will create Distributions for data in R. As always, we will
consistently use the [**Project Mosaic**](https://mosaic-web.org)
ecosystem of packages in R (`mosaic`, `mosaicData` and `ggformula`).

::: {.callout-tip}
Note the standard method for all commands from the `mosaic` package:
`goal( y ~ x | z, data = mydata, …)`

With `ggformula`, one can create any graph/chart using:
`gf_geometry(y ~ x | z, data = mydata)` OR
`mydata %>% gf_geometry( y ~ x | z)`

The second method may be preferable, especially if you have done some
data manipulation first! More later!
:::

## {{< iconify pajamas issue-type-test-case >}} Case Study #1: Galton Dataset from `mosaicData`

Let us inspect what datasets are available in the package `mosaicData`.
Type `data(package = "mosaicData")` in your Console to see what datasets
are available.

Let us choose the famous `Galton` dataset:

```{r}
#| label: Chunk-1
#| warning: false
#| message: false
#| echo: true
data("Galton")
Galton_inspect <- inspect(Galton)
Galton_inspect

```

The data is described as:

> A data frame with 898 observations on the following variables.
>
> -   `family` a factor with levels for each family
> -   `father` the father's height (in inches)
> -   `mother` the mother's height (in inches)
> -   `sex` the child's sex: F or M
> -   `height` the child's height as an adult (in inches)
> -   `nkids` the number of adult children in the family, or, at least,
>     the number whose heights Galton recorded.

There is a lot of Description generated by the `mosaic::inspect()`
command ! What can we say about the dataset and its variables? How big
is the dataset? How many variables? What types are they, Quant or Qual?
If they are Qual, what are the *levels*? Are they *ordered* levels?
Discuss!

### {{< iconify eos-icons counting >}} Counts, and Charts with Counts

Now that we know the variables, let us look at **counts** of data
observations(rows). We know from our examination of variable types that
counting of observations must be done on the basis of **Qualitative**
variables. So let us count and plot the counts in bar charts.

::: {.callout-note title="Question"}
Q.1 How many families in the data for each value of `nkids`(i.e. Count
of families by size)?

```{r}
#| layout-ncol: 2
Galton_counts <- Galton %>%
  group_by(nkids) %>% 
  summarise(children = n()) %>% 
  # just to check
  mutate(
    No_of_families = as.integer(children/nkids),
    # Why do we divide
    
    running_count_of_children = cumsum(children),
    running_count_of_families = cumsum(No_of_families)) 
Galton_counts

Galton_counts %>% 
  gf_col(No_of_families ~ nkids) %>% 
  gf_theme(theme_classic())

```

**Insight**: There are 32 1-kid families; and $128/8 = 16$ 8-kid
families! There is one great great 15-kid family. (Did you get the idea
behind why we divide here?)
:::

::: {.callout-note title="Question"}
Q.2. What is the count of Children by `sex` of the child and by family
size `nkids`?

```{r}
#| layout-ncol: 2
#| message: false
#| warning: false
Galton_counts_by_sex <- Galton %>% 
  mutate(family = as.integer(family)) %>% 
  group_by(nkids, sex) %>% 
  summarise(count_by_sex = n()) %>% 
  ungroup() %>% 
  group_by(sex) 
Galton_counts_by_sex

Galton_counts_by_sex %>% 
  gf_col(count_by_sex ~ nkids | sex, fill = ~sex) %>% 
  gf_theme(theme_classic())

```

**Insight**: Hmm...decent gender balance overall, across family sizes
`nkids`.
:::

::: {.callout-note title="Follow-up Question"}
Follow up Question: How would we look for "gender balance" in individual
families? Should we look at the `family` column ?

```{r}
#| message: false
#| warning: false
#| layout-ncol: 2

Galton_gender_family <- Galton %>% 
  mutate(family = as.integer(family)) %>% 
  group_by(family, sex) %>% 
  summarise(count_by_sex = n()) %>% 
  ungroup() %>% 
  group_by(sex)
Galton_gender_family
Galton_gender_family %>% 
  gf_col(count_by_sex ~ family | sex, fill = ~ sex) %>% 
  gf_theme(theme_minimal())
  
  
```

**Insight**: The No of Children were distributed similarly across family
size`nkids`... However, this plot is too crowded and does not lead to
any great insight. Using `family` ID was silly to plot against, wasn't
it? Not all exploratory plots will be "necessary" in the end. But they
are part of the journey of getting better acquainted with the data!
:::

### {{< iconify icon-park-outline overall-reduction >}} {{< icon-park-outline chart-histogram >}} Stat Summaries and Distributions

OK, on to the Quantitative variables now! What Questions might we have,
that could relate *not to counts by Qual variables*, but to the numbers
in Quant variables. Stat measures, like their ranges, max and min?
Means, medians, distributions? And how these vary on the basis of *Qual*
variables? All this using `histograms` and `densities`.

::: {.callout-note title="Summary Stats"}
As Stigler[@stigler2016] said, **summaries** are the first thing to look
at in data. `skimr::skim` has already given us a lot summary data for
Quant variables. We can now use `mosaic::favstats` to develop these
further, by **slicing / facetting** these wrt other Qual variables. Let
us tabulate some quick stat summaries of the important variables in
`Galton`.

```{r}

# summaries facetted by sex of child
Galton_measures <- favstats(~ height | sex, data = Galton)
Galton_measures

```

**Insight**: We saw earlier that the mean `height` of the Children was
66 inches. However, are Sons taller than Daughters? Difference in mean
`height` is 5 inches! AND...that was the same difference between
`fathers` and `mothers` mean heights! Is it so simple then?
:::

::: {.callout-note title="Question"}
Q.4 How are the heights of the children distributed? Here is where we
need a `gf_histogram`...

```{r}
Galton %>% 
  gf_histogram(~ height,bins = 30) %>% # ALWAYS try several settings for "bins"
  gf_vline(xintercept = mean(Galton$height), color = "red") %>% 
  gf_label(label = glue::glue("Mean Value\n {mean(Galton$height)}"),
           gformula = 85 ~ 70) %>% # Where do we want the label (y ~ x) %>% 
  gf_labs(title = "Galton Dataset",
          subtitle = "",
          x = "Heights of Children",
          y = "Counts",
          caption = "Using `ggformula` and Classic Theme") %>% 
  gf_theme(theme_classic())

```

**Insight**: Fairly symmetric distribution...but there are a few very
short and some very tall children! *Always try to change the no. of bins
to check of we are missing some pattern.*
:::

::: {.callout-note title="Question"}
Q.5 Is there a difference in height distributions between Male and
Female children?(*Quant variable sliced by Qual variable*)

```{r}
mean_daughters <- Galton_measures %>% 
             filter(sex == "F") %>% select(mean) %>% as.numeric()
mean_sons <- Galton_measures %>% 
             filter(sex == "M") %>% select(mean) %>% as.numeric()
Galton %>% 
  gf_density(~ height, group = ~sex, fill = ~ sex, alpha = 0.5) %>% 
  gf_vline(xintercept = mean_daughters) %>%
  gf_vline(xintercept = mean_sons) %>% 
  
   gf_label(label = glue::glue("Mean Daughters' heights\n {mean_daughters}"),
           gformula = 0.15 ~ 59, # Where do we want the label (y ~ x) 
           show.legend = FALSE,
           fill = "white") %>% 
  
   gf_label(label = glue::glue("Mean Sons' heights \n {mean_sons}"),
           gformula = 0.15 ~ 75,  # Where do we want the label (y ~ x) 
           show.legend = FALSE,
           fill = "white") %>% 
  gf_theme(theme_classic())

```

**Insight**: There is a visible difference in average heights between
girls and boys. Is that significant, however? We will need a statistical
inference test to figure that out!! Claus Wilke[^1] says comparisons of
Quant variables across groups are best made between densities and not
histograms...
:::

::: {.callout-note title="Question"}
Q.6 Are Mothers generally shorter than fathers?

```{r}
mean_fathers <- Galton %>% mosaic::mean(~ father, data = .) %>% as.numeric()

mean_mothers <- Galton%>% mean(~ mother, data = .) %>% as.numeric()

Galton %>% 
  gf_density(~ father, fill = "dodgerblue", alpha = 0.3) %>% 
  gf_density(~ mother, fill = "red", alpha = 0.3) %>% 
    gf_vline(xintercept = mean_mothers) %>%
  gf_vline(xintercept = mean_fathers) %>% 
  gf_labs(x = "Parents' Heights") %>% 
    
   gf_label(label = glue::glue(" Average Mother \n {mean_mothers}"),
           gformula = 0.15 ~ 60, # Where do we want the label (y ~ x) 
           show.legend = FALSE,
           fill = "white") %>% 
  
   gf_label(label = glue::glue("Average Father\n {mean_fathers}"),
           gformula = 0.15 ~ 75,  # Where do we want the label (y ~ x) 
           show.legend = FALSE,
           fill = "white") %>% 
  
  gf_theme(theme_classic())


```

**Insight**: Yes, moms are on average shorter than dads in this dataset.
Again, is this difference statistically significant? We will find out in
when we do **Inference**.
:::

::: {.callout-note title="Question"}
Q.7a. Are heights of children different based on the number of kids in
the family? And For Male and Female children?

```{r}
# Boxplot series for daughters
gf_boxplot(height ~ nkids, group = ~ nkids, fill = ~ sex, 
           data = Galton %>% filter(sex == "F")) %>% 
# Boxplot series for sons
  gf_boxplot(height ~ nkids, group = ~ nkids, fill = ~ sex, 
             data = Galton %>% filter(sex == "M")) %>% 
  gf_labs(title = "Daughters vs Sons: Who is Taller?",
          subtitle = "Family Size does not change height disparity ;-D",
          x = "Family Size", y = "Heights of Children",
          caption = "Made with `ggformula` and Classic Theme") %>% 
  
  gf_theme(theme_classic())


```

**Insight**: So, at all family "strengths" (`nkids`), the sons are
taller than the daughters, based on distributions. Box plots are used to
show distributions of numeric data values and compare them between
multiple groups (i.e Categorical Data, here `sex` and `nkids`).
:::

::: {.callout-note title="Follow-up Question"}
Q. 8a. Is height difference between sons and daughters related to height
difference between `father` and `mother`?

Differences between `father` and `mother` heights influencing
`height`...this would be like `height ~ (father-mother)`. This would be
a relationship between **two** Quant variables. A histogram would not
serve here and we plot this as a **Scatter Plot**:

```{r}
#| layout: [[40], [60]]
Galton_parents_diff <- Galton %>% 
  group_by(family,sex) %>% 
  
  # Parental Height Difference
  mutate(diff_height = father - mother) %>% 
  select(family, sex, height, diff_height) %>% 
  ungroup() %>% 
  group_by(sex)
Galton_parents_diff

Galton_parents_diff %>% 
  gf_point(height ~ diff_height, color = ~ sex, data = Galton_parents_diff) %>%
  gf_lims(y = c(55,82)) %>% # note: Y-AXIS does not go to zero!!
  
  gf_rect(55 + 80 ~ 0 + 18, fill = "grey80", 
          color = "white", alpha = 0.02) %>%
  gf_rect(55 + 80 ~ -5 + 0, fill = "moccasin", 
          color = "white", alpha = 0.02) %>%   
  
  # Note the repetition!
  gf_point(height ~ diff_height, color = ~ sex, data = Galton_parents_diff) %>%
  gf_smooth(method = "lm", se = FALSE) %>% 
  gf_labs(x = "Parental Height Difference",
          y = "Children's Heights") %>% 
  
  gf_label(label = glue::glue("Taller Mothers"),
           gformula = 75 ~ -2.5, # Where do we want the label (y ~ x) 
           show.legend = FALSE,
           fill = "white") %>% 
  
  gf_label(label = glue::glue("Taller Fathers"),
           gformula = 75 ~ 15,  # Where do we want the label (y ~ x) 
           show.legend = FALSE,
           fill = "white") %>% 
  
  gf_vline(xintercept = 0, linewidth = 2, 
           title = "Does Parental Height Difference affect 
                    Children's Height?",
           caption = "Note Y-axis does not go to zero") %>% 
  
  gf_theme(theme_classic())

```

**Insight**: There seems no relationship, or a very small one, between
children's `height`s on the y-axis and the difference in parental height
differences (`father - mother`) on the x-axis...
:::

And so on.....we can proceed from simple visualizations based on
Questions to larger questions that demand **inference** and
**modelling**. We hinted briefly on these in the above Case Study.

## {{< iconify pajamas issue-type-test-case >}} Case Study-2: Dataset from `NHANES`

Let us try the `NHANES` dataset. [Try `help(NHANES)` in your
Console.]{.aside}

```{r}
data("NHANES")
```

### {{< iconify grommet-icons inspect >}} Look at the Data

```{r}
#| column: body-outset-right
skimr::skim(NHANES)
```

Again, lots of data from `skim`, about the Quant and Qual variables.
Spend a little time looking through this output.

-   Which variables could have been data that was *given/stated* by each
    respondent?
-   And which ones could have been *measured* dependent data variables?
    Why do you think so?
-   Why is there so much *missing* data? Which variable are the most
    affected by this?\

### {{< iconify eos-icons counting >}} Counts, and Charts with Counts

::: {.callout-note title="Question"}
Q.1 What are the Education levels and the counts of people with those
levels?

```{r}

NHANES %>% 
  group_by(Education) %>% 
  summarise(total = n())

# This also works
# tally(~Education, data = NHANES) %>% as_tibble()

```

**Insight**: The count goes up as we go from lower Education levels to
higher. Need to keep that in mind. How do we understand the large number
of `NA` entries? What could have led to these entries?
:::

::: {.callout-note title="Question"}
Q.2 How do counts of Education vs Work-status look like?

```{r}
#| message: false
#| layout: [[30], [70]]
NHANES %>% 
  mutate(Education = as.factor(Education)) %>% 
  group_by(Work,Education) %>% 
  summarise(count = n())

NHANES %>% 
  group_by(Work, Education) %>% 
  summarise(count = n()) %>% 
  gf_col(count ~ Education, fill = ~ Work, position = "dodge") %>% 
  gf_theme(theme_classic())

```

**Insight**: Clear increase in the number of Working people as Education
goes from 8th Grade to College. No surprise. Are the `NotWorking` counts
a surprise?
:::

### {{< iconify icon-park-outline overall-reduction >}} {{< icon-park-outline chart-histogram >}} Stat Summaries, Histograms, and Densities

::: {.callout-note title="Question"}
Q.3. What is the distribution of Physical Activity Days, across Gender?
Across Education?

```{r}
#| layout-ncol: 2
#| column: body-outset-right
#| warning: false

NHANES %>% 
  gf_histogram(~ PhysActiveDays | Gender, fill = ~Gender) %>% 
  gf_theme(theme_classic())

NHANES %>% 
  gf_histogram(~ PhysActiveDays | Education, fill = ~ Education) %>% 
  gf_theme(theme_classic())

```

**Insight**: Can we conclude anything here? The **populations** in each
category are different, as indicated by the different heights of the
bars, so what do we need to do? Take percentages or ratios of course,
**per-capita**! How would one do that?
:::

::: {.callout-note title="Question"}
Q.3a. What is the distribution of Physical Activity Days, across
Education and Sex, **per capita**?

```{r}
#| layout-ncol: 2
#| message: false
NHANES %>% 
  group_by(Gender) %>% 
  summarize(mean_active = mean(PhysActiveDays,na.rm = TRUE))

NHANES %>% 
  group_by(Education, Work) %>% 
  summarize(mean_active = mean(PhysActiveDays,na.rm = TRUE))

```

**Insight**: Hmm..no great differences in per-capita physical activity.
Females are marginally more active than males. No need to even plot
this.
:::

::: {.callout-note title="Question"}
Q.4. How are people's Ages distributed across levels of Education?

```{r}
#| warning: false
#| layout-nrow: 1
# Recall there are missing data
gf_boxplot(Age ~ Education,
           fill = ~ Education, # Always a good idea to fill boxes
           data = NHANES) %>%
  gf_theme(theme_classic()) %>% 
  
  # And to turn this into an interactive plot
  plotly::ggplotly()

```

**Insight**: Older age groups are somewhat more heavily represented in
groups with lower educational status. College Graduates also have
slightly narrow age....That is a nice Question for some Inferential
Modelling. And how to interpret the NA group?
:::

::: {.callout-note title="Question"}
Q.5. How is Education distributed over Race?

```{r}
#| warning: false
#| message: false
#| layout: [[25], [75]]

NHANES_by_Race1 <- NHANES %>% 
  group_by(Race1) %>% 
  summarize(population = n())
NHANES_by_Race1

NHANES %>% group_by(Education, Race1) %>% 
  summarize( n = n()) %>% 
  left_join(NHANES_by_Race1, by = c("Race1" = "Race1")) %>% 
  mutate(percapita_educated = (n/population)*100) %>% 
  ungroup() %>%  
  group_by(Race1) %>%
  gf_col(percapita_educated ~ Education | Race1, fill = ~ Race1) %>%
  gf_refine(coord_flip()) %>%  
  gf_theme(theme_classic()) %>% 

  # And to turn this into an interactive plot
  plotly::ggplotly()


```

**Insight**: Blacks, Hispanics, and Mexicans tend to have fewer people
with college degrees, as a percentage of their population. Asians and
other immigrants have a significant tendency towards higher education!
:::

::: {.callout-note title="Question"}
Q.6. What is the distribution of people's BMI, split by Gender? By
Race1?

```{r}
#| warning: false
#| message: false
#| layout-ncol: 2
# One can also plot both histograms and densities in an overlay fashion,

NHANES %>% 
  group_by(Gender) %>% 
  gf_density(~ BMI | Gender,fill = ~ Gender , alpha = 0.4) %>% 
  gf_fitdistr(dist = "dnorm",color =  ~ Gender) %>% 
  gf_theme(theme_classic()) %>% 

  # And to turn this into an interactive plot
  plotly::ggplotly()

NHANES %>% group_by(Race1) %>% 
  gf_density(~ BMI | Race1, fill = ~ Race1) %>% 
  gf_fitdistr(dist = "dnorm") %>% 
  gf_theme(theme_classic()) %>% 

  # And to turn this into an interactive plot
  plotly::ggplotly()
```

**Insight**: Blacks tend to have larger portions of their populations
with larger BMI. So these races perhaps tend to obesity. By and large
BMI distributions are normal.
:::

::: {.callout-note title="Question"}
Q.7. What is the distribution of people's Testosterone level vs BMI?
Split By Race1?

```{r}
#| warning: false
NHANES %>%  
  gf_density_2d(Testosterone ~ BMI | Race1) %>%
  gf_theme(theme_classic()) %>% 
plotly::ggplotly()

```

**Insight**: Low testosterone levels exist across all BMI values, but
healthy levels of T exists only over a smaller range of BMI.
:::

## {{< iconify pajamas issue-type-test-case >}} Case Study#3: A complete example

Here is a dataset from Jeremy Singer-Vine's blog, [Data Is
Plural](https://www.data-is-plural.com/). This is a list of all books
banned in schools across the US.

{{< downloadthis ../data/banned.csv dname="banned.csv" label="Download the data"banned books icon="database-fill-down" type="info" >}}

```{r}
#| label: Banned Books
banned <- readxl::read_xlsx(path = "../data/banned.xlsx",
                            sheet = "Sorted by Author & Title")
banned
names(banned)

```

Clearly the variables are *all* Qualitative, except perhaps for *Date of
Challenge/Removal, (which in this case has been badly mangled by Excel)
So we need to make **counts** based on the* levels\* of the Qual
variables and plot Bar/Column charts.

Let us quickly make some Stat Summaries ( using `inspect`)

```{r}
mosaic::inspect(banned)

```

**Insight**: Clearly the variables are *all* Qualitative, except perhaps
for *Date of Challenge/Removal*, (which in this case has been badly
mangled by Excel). So we need to make **counts** based on the\* levels\*
of the Qual variables and plot Bar/Column charts. We will not find much
use for histograms or densities.

Let us try to answer this question:

::: {.callout-note title="Question"}
Q.1.What is the count of banned books by type? By US state?

```{r}
#| message: false
#| layout: [[60],[40]]
banned %>% count(`Type of Ban`, sort = TRUE)
banned %>% count(State, sort = TRUE)

```

**Insight**: There are four types of book bans, and
`Banned Pending Investigation` is the most frequent, but other types of
bans are frequent too. Texas is the leader in book bans!

From: <https://en.wikipedia.org/wiki/Bible_Belt>

> The Bible Belt is a region of the Southern United States in which
> socially conservative Protestant Christianity plays a strong role in
> society. Church attendance across the denominations is generally
> higher than the nation's average. The region contrasts with the
> religiously diverse Midwest and Great Lakes, and the Mormon corridor
> in Utah and southern Idaho.

![Bible
Belt](https://upload.wikimedia.org/wikipedia/commons/c/c2/BibleBelt.png){fig-align="center"}
:::

::: {.callout-note title="Question"}
Q.2. What is the count of banned books by type **and** by US state?\*

```{r}
#| message: false

banned %>% group_by(State, `Type of Ban`) %>% 
  count(State, sort = TRUE) %>% 
  slice_max(n = 10, order_by = n, with_ties = TRUE) %>% 
  gf_col(reorder(State, n) ~ n, fill = ~`Type of Ban`,
         xlab = "No of Banned Books", ylab = "State") %>% 
  gf_refine(scale_fill_brewer(palette = "Set2")) %>% 
  gf_theme(theme_classic())

```

**Insight**: Clearly Texas is the leading state in book bans in schools.
California clearly doesn't bother itself with these things!!
:::

## {{< iconify fluent-mdl2 decision-solid >}} Conclusion

And that is a wrap!! Try to work with this procedure:\

-   Inspect the data using `skim` or `inspect`
-   Identify Qualitative and Quantitative variables
-   Notice variables that have missing data and decide if that matters
-   Develop Counts of Observations for combinations of Qualitative
    variables (factors)
-   Develop Histograms and Densities, and slice them by Qualitative
    variables to develop faceted plots as needed
-   Continue with other *Descriptive Graphs* as needed
-   At each step **record the insight and additional questions!!**
-   Onwards with Correlations and other visuals...
-   And then on the *inference* and *modelling*!!

## {{< iconify ooui references-rtl >}} References

1.  A detailed analysis of the NHANES dataset,
    <https://awagaman.people.amherst.edu/stat230/Stat230CodeCompilationExampleCodeUsingNHANES.pdf>

[^1]: Fundamentals of Data Visualization (clauswilke.com)
