---
title: "\U0001F0CF Permutation Test for Two Proportions"
author: "Arvind Venkatadri"
date: 10/Nov/2022
date-modified: "`r Sys.Date()`"
abstract: "Using Permutation Tests to check the equivalence of two proportions"
order: 80
categories:
- Permutation
- Monte Carlo Simulation
- Random Number Generation
- Distributions
- Generating Parallel Worlds
---

## {{< fa folder-open >}} Slides and Tutorials

| <a href="./files/two-props.qmd"><i class="fa-brands                     
                     fa-r-project"></i> R Tutorial</a>                    | <a href="./files/two-props.ows"> <iconify-icon icon="fluent-emoji:orange-circle"></iconify-icon> Orange Tutorial</a> | <a href="./files/two-props.rda"> <i class="fa-solid fa-person-rays"></i> Radiant Tutorial</a>  | <a href="./files/data/sim-data.zip"> <i class="fa-solid fa-database"></i> Datasets</a> |
|-------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------|

## {{< iconify noto-v1 package >}} Setting up R packages

```{r}
#| label: setup
#| include: true
#| message: false
#| warning: false
knitr::opts_chunk$set(echo = TRUE,message = TRUE,warning = TRUE, fig.align = "center")
options(digits=2)
library(tidyverse)
library(mosaic)
library(ggmosaic) # plotting mosaic plots for Categorical Data
library(vcd) # Creating Tables and plotting mosaic charts

### Dataset from Chihara and Hesterberg's book (Second Edition)
library(resampledata)

```

## Introduction

We saw from the diagram created by Allen Downey that *there is only one
test*! We will now use this philosophy to develop a technique that
allows us to mechanize several *Statistical Models* in that way, with
nearly identical code.

We will use two packages in R, `mosaic` to develop our intuition for
what are called **permutation** based statistical tests. (There is also
a more recent package called `infer` in R which can do pretty much all
of this, including visualization. In my opinion, the code is a little
too high-level and does not offer quite the detailed insight that the
`mosaic` package does).

## Testing for Two or More Proportions

Let us try a dataset with Qualitative / Categorical data. This is the
`General Social Survey GSS dataset`, and we have people with different
levels of `Education` stating their opinion on the `Death Penalty`. We
want to know if these two Categorical variables have a correlation, i.e.
can the opinions in favour of the `Death Penalty` be explained by the
`Education` level?

Since data is Categorical ( both variables ), we need to take `counts`
in a table, and then implement a `chi-square test`. In the test, we will
permute the `Education` variable to see if we can see how significant
its *effect size* is.

```{r}

data(GSS2002)
inspect(GSS2002)

```

Note how *all* variables are Categorical !! `Education` has five
`levels`, and of course `DeathPenalty` has three:

```{r}

GSS2002 %>% count(Education)
GSS2002 %>% count(DeathPenalty)

```

Let us drop NA entries in `Education` and `Death Penalty` and set up a
**Contingency Table**.

```{r,warning=FALSE,message=FALSE}


gss2002 <- GSS2002 %>% 
  dplyr::select(Education, DeathPenalty) %>% 
  tidyr::drop_na(., c(Education, DeathPenalty))



gss_table <- tally(DeathPenalty ~ Education, data = gss2002)
gss_table %>% 
  addmargins()

```

## Contingency Table Plots

The Contingency Table can be plotted, as we have seen, using a `mosaic`
plot using several packages:

::: panel-tabset
### Using `ggformula`

Need a little more work, to convert the Contigency Table into a tibble:

```{r warning=FALSE,message=FALSE}
# https://stackoverflow.com/questions/19233365/how-to-create-a-marimekko-mosaic-plot-in-ggplot2

gss_summary <- gss2002 %>%
  mutate(
    Education = factor(
      Education,
      levels = c("Bachelors", "Graduate", "Jr Col", "HS", "Left HS"),
      labels = c("Bachelors", "Graduate", "Jr Col", "HS", "Left HS")
    ),
    DeathPenalty = as.factor(DeathPenalty)
  ) %>%
  group_by(Education, DeathPenalty) %>%
  summarise(count = n()) %>% # This is good for a chisq test
  
  # Add two more columns to facilitate mosaic/Marrimekko Plot
  # 
  mutate(edu_count = sum(count), 
         edu_prop = count / sum(count)) %>%
  ungroup() 

gf_col(edu_prop ~ Education, data = gss_summary,
       width = ~ edu_count, 
       fill = ~ DeathPenalty,
       stat = "identity", 
       position = "fill", 
       color = "black") %>% 
  
  gf_text(edu_prop ~ Education, 
          label = ~ scales::percent(edu_prop),
          position = position_stack(vjust = 0.5)) %>% 
  
  gf_facet_grid(~ Education, 
                scales = "free_x", 
                space = "free_x") %>% 
  
  gf_theme(scale_fill_manual(values = c("orangered", "palegreen3"))) %>% 
  gf_theme(theme_void())
  

 

```

### Using `vcd`

```{r}
vcd::mosaic(gss_table, gp = shading_hsv)

```

### Using `ggmosaic`

```{r mosaic-plot,warning=FALSE}
#library(ggmosaic)

ggplot(data = gss2002) +
  geom_mosaic(aes(x = product(DeathPenalty, Education), 
                  fill = DeathPenalty))

```
:::

## Observed Statistic: the $X^2$ metric

When there are multiple proportions involved, the $X^2$ test is what is
used.

::: panel-tabset
### Intuitive Explanation

Let us look at the Contingency Table that we have:

```{r echo=FALSE}

gss_table %>% addmargins() %>% 
  kableExtra::kbl(caption = "Contigency Table") %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")

```

In the chi-square test, we check whether the two ( or more ) categorical
variables are independent. To do this we perform a simple check on the
Contingency Table. We first *re-compute* the totals in each row and
column, based on what we could **expect** if there was independence
(NULL Hypothesis). If the two variables were independent, then there
should be **no difference** between real and expected scores.

How do we know what scores to expect?

Consider the entry in location (1,1): 117. The number of **expected**
entries there is probability of an entry landing in that square times
the total number of entries:

::: column-body-outset
```{=tex}
\begin{align}

\text{Expected Value at location[1,1]}
&= p_{row_1} * p_{col_1} * \text{Total Scores}\\\
&= \frac{\text{Row-1-Total}}{\text{Total Scores}} * \frac{\text{Col-1-Total}}{\text{Total Scores}} * \text{Total Scores}\\\
&= \frac{898}{1307} * \frac{189}{1307} * 1307\\\
&= 130


\end{align}
```
:::

Proceeding in this way for all the 15 entries in the Contingency Table,
we get the "Expected" Contingency Table. Here are both tables for
comparison:

```{r echo=FALSE}
# using prop.table with appropriate margin argument gives us a slick matrix of probabilities

gss_exp <- gss_table * gss_table / ((
  prop.table(gss_table, margin = 2) * prop.table(gss_table, margin = 1)
) * 1307)

```

```{r echo=FALSE}

gss_exp %>% addmargins() %>% 
   kableExtra::kbl(caption = "Expected Contigency Table", digits = 0) %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")

```

```{r echo=FALSE}

gss_table %>% addmargins() %>% 
   kableExtra::kbl(caption = "Actual Contigency Table") %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")


```

The $X^2$ statistic is sum of squared differences between `Observed` and
`Expected` scores, scaled by the `Expected Scores`. For location \[1,1\]
this would be: $(117-130)^2/189$. Do try to compute all of these and the
$X^2$ statistic by hand !!

### Code

Let us now perform the base `chisq test`: We need a `table` and then the
`chisq` test:

```{r}

# gss_table <- tally(DeathPenalty ~ Education, data = gss2002)
# gss_table

# Get the observed chi-square statistic
observedChi2 <- mosaic::chisq(tally(DeathPenalty ~ Education, data = gss2002))
observedChi2

# Actual chi-square test
stats::chisq.test(tally(DeathPenalty ~ Education, data = gss2002))

```

We see that our observed $X^2 = 23.45$.
:::

### Hypotheses Definition

What would our Hypotheses be?

$H_0: \text{Education does not affect votes for Death Penalty}$
$H_a: \text{Education affects votes for Death Penalty}$

### Permutation Test for `Education`

We should now repeat the test with permutations on `Education`:

```{r}

null_chisq <- do(10000) * 
  chisq.test(tally(DeathPenalty ~ shuffle(Education), 
                   data = gss2002))

head(null_chisq)

gf_histogram( ~ X.squared, data = null_chisq) %>% 
  
  gf_vline(xintercept = observedChi2, 
           color = "red") %>% 
  gf_theme(theme = theme_classic())

prop1(~ X.squared >= observedChi2, data = null_chisq)

```

The `p-value` is well below our threshold of $0.05$, so we would
conclude that `Education` has a significant effect on `DeathPenalty`
opinion!

## Conclusion

Why would a permutation test be a good idea here?

In our basic $X^2$ test, we calculate the test statistic of $X^2$ and
look up a *theoretical* null distribution for that statistic, and see
how unlikely our observed value is.

With a permutation test, there are *no assumptions* of the null
distribution: this is computed based on real data. We note in passing
that, in this case, since the number of `cases` in each cell of the
Contingency Table are fairly high ( \>= 5) the resulting NULL
distribution is of the $X^2$ variety.

## References
